---
title: "R Notebook"
output: html_notebook
---

```{r}
easypackages::libraries(c("readr","readxl","dplyr","tidyr","ggplot2","ggpubr","cowplot","grid","ggplotify","uwot","purrr","vegan","cluster"))
# insert path to folders
directory_path <- "" # CHANGE THIS to path to director enclosing code and data sets
folder_path <- paste0(directory_path, "/Supplemental Tables/") # folder containing downloaded data sets
intermediate_folder_path <- paste0(directory_path, "/Intermediate Files Used in Plots/") # path containing intermediate outputs
folder_path_save <- paste0(directory_path, "/Complete Figures/")# define folder path to save images to

# Load Supplemental Data
combined_metadata_dist <- read_csv(paste0(folder_path,"Data Table S1.csv"))
all_zs_combined <- read_csv(paste0(folder_path,"Data Table S2.csv"))
all_hits_combined <- read_csv(paste0(folder_path,"Data Table S3.csv"))
Viral_Annotations <- read_csv(paste0(folder_path,"Data Table S4.csv"))
AVARS_virscores <- read_csv(paste0(folder_path,"Data Table S5.csv"))
IgG_Data <- read_csv(paste0(folder_path,"Data Table S6.csv"))

# Load intermediate files
adult_viral_peptide_prevalence <- read.csv(paste0(intermediate_folder_path,"adult_viral_peptide_prevalence.csv"))
Child_viral_peptide_prevalence <- read.csv(paste0(intermediate_folder_path,"Child_viral_peptide_prevalence.csv"))

# Load required functions
subset_frame_fxn<-function(wide_df,sample_metadata,dataset_lab) {
  sample_list<-sample_metadata$identifier[which(sample_metadata[[dataset_lab]]==TRUE)]
  filtered_df<-wide_df %>% select(c(id,all_of(sample_list)))
  return(filtered_df)
}
family_breakdown_prop_fn<-function(Hit_frame,peptide_family_labels) {
  peptide_labels <- peptide_family_labels %>% select(c(id,family_label))
  family_breakdown <- Hit_frame %>% left_join(peptide_labels) %>% pivot_longer(-c(id,family_label),names_to = "identifier",values_to = "Hit") %>% filter(Hit==1)%>% group_by(identifier,family_label) %>% summarise(n_hits=sum(Hit)) %>% pivot_wider(id_cols = identifier,names_from = family_label,values_from = n_hits,values_fill = 0) %>% mutate(tot_hits=Other+Herpesviridae+ Picornaviridae+ Orthomyxoviridae+ Pneumoviridae+Adenoviridae+Coronaviridae+Paramyxoviridae) 
  family_breakdown_prop<-family_breakdown %>% mutate(Other=Other/tot_hits,Herpesviridae=Herpesviridae/tot_hits, Picornaviridae=Picornaviridae/tot_hits, Orthomyxoviridae=Orthomyxoviridae/tot_hits, Pneumoviridae=Pneumoviridae/tot_hits, Adenoviridae=Adenoviridae/tot_hits,  Coronaviridae=Coronaviridae/tot_hits, Paramyxoviridae=Paramyxoviridae/tot_hits)
return(family_breakdown_prop)
}
pal1<-c("#F87E2E", "#FFEC33", "#1FAA79", "#14C4F8", "#0077C8", "#9870F5", "#D53EE1", "#FF36B0", "#D72638", "#6E6E6E") 
```



### Figure 2A
# Create UMAP
Visualize the UMAP 2D results: color by time intervals  
```{r visualize the UMAP calculate with hits, fig.height = 3, fig.width = 5}
adult_child_longzs <- all_hits_combined %>%
  dplyr::select(c(id, all_of(IgG_Data$identifier))) %>%
  tidyr::pivot_longer(-id, names_to = "identifier", values_to = "hit") %>%
  dplyr::left_join(
    IgG_Data %>% dplyr::select(c(identifier, donor, time_interval, Age_Years, age_group, interval_label)),
    by = "identifier"
  ) %>%
  dplyr::mutate(age_group = ifelse(age_group == ">1yr", "1-3yr", age_group)) %>%
  dplyr::bind_rows(
    subset_frame_fxn(all_hits_combined, combined_metadata_dist, "adult_cross") %>%
      tidyr::pivot_longer(-id, names_to = "identifier", values_to = "hit") %>%
      dplyr::left_join(
        combined_metadata_dist %>%
          dplyr::mutate(time_interval = 40, age_group = "adult", interval_label = "Adult") %>%
          dplyr::select(c(identifier, donor, time_interval, Age_Years, age_group, interval_label)),
        by = "identifier"
      )
  )

temp <- adult_child_longzs %>%
  dplyr::select(c(identifier, interval_label)) %>%
  dplyr::distinct() %>%
  dplyr::group_by(interval_label) %>%
  dplyr::summarise(n_in_group = dplyr::n(), .groups = "drop") %>%
  dplyr::mutate(interval_label_updated = paste0(interval_label, " (", n_in_group, ")"))

adult_child_longzs <- adult_child_longzs %>%
  dplyr::left_join(temp %>% dplyr::select(-n_in_group), by = "interval_label")
rm(temp)

# Sample metadata (one row per identifier)
sample_meta <- adult_child_longzs %>%
  dplyr::group_by(donor, identifier, Age_Years, time_interval, interval_label, interval_label_updated) %>%
  dplyr::summarise(sum_hit = sum(hit), .groups = "drop") %>%
  dplyr::mutate(key = identifier)

# Build sample-by-peptide binary matrix (rows = identifier)
hit_unfiltered_matrix <- adult_child_longzs %>%
  dplyr::mutate(key = identifier) %>%
  tidyr::pivot_wider(id_cols = key, names_from = id, values_from = hit, values_fill = 0) %>%
  tibble::column_to_rownames(var = "key") %>%
  as.matrix()

# Drop peptides with extreme prevalence (>=5 and <= N-5)
cs <- colSums(hit_unfiltered_matrix)
keep <- cs >= 5 & cs <= (nrow(hit_unfiltered_matrix) - 5)
hit_unfiltered_matrix <- hit_unfiltered_matrix[, keep, drop = FALSE]
rm(cs, keep)

# Jaccard UMAP for uwot (workaround for metric limitations) 
# Build exact kNN from a distance matrix (avoids Annoy 'precomputed' issue)
knn_from_dist <- function(D, k) {
  n <- nrow(D)
  k <- min(k, n - 1)

  idx <- t(vapply(seq_len(n), function(i) {
    ord <- order(D[i, ], na.last = NA)
    ord <- ord[ord != i]        # drop self
    ord[seq_len(k)]
  }, integer(k)))

  dist <- t(vapply(seq_len(n), function(i) {
    D[i, idx[i, ]]
  }, numeric(k)))

  list(idx = idx, dist = dist)
}

# Compute Jaccard distances once
D_jacc <- as.matrix(vegan::vegdist(hit_unfiltered_matrix, method = "jaccard", binary = TRUE))
storage.mode(D_jacc) <- "double"

# Precompute neighbors for the desired n_neighbors
n_neighbors <- 15
nn_jacc <- knn_from_dist(D_jacc, k = n_neighbors)

# Run uwot using the precomputed neighbor graph
set.seed(10)
hit_unfiltered_umap_layout <- uwot::umap(
  X = hit_unfiltered_matrix,
  n_neighbors = n_neighbors,
  n_components = 2,
  nn_method = nn_jacc,   # <- key workaround
  n_epochs = 200,
  init = "spectral",
  min_dist = 0.1,
  verbose = FALSE
)

# Format for plotting
hit_unfiltered_umap_dt <- as.data.frame(hit_unfiltered_umap_layout) %>%
  tibble::rownames_to_column(var = "key") %>%
  dplyr::left_join(sample_meta, by = "key")

Fig2A <- ggplot(
  hit_unfiltered_umap_dt,
  aes(
    x = V1, y = V2,
    color = factor(
      interval_label_updated,
      levels = c(
        "birth (25)", "3mo (22)", "6mo (23)", "1yr (25)",
        "1.5yr (22)", "2yr (20)", "3yr (23)", "Adult (559)"
      )
    )
  )
) +
  geom_point(size = 1.5, alpha = 0.5) +
  stat_ellipse(level = 0.95, linetype = "longdash", type = "t") +
  scale_color_manual(name = "Age group", values = pal1[c(3,4,5,7,8,9,1,10)]) +
  theme_bw() +
  theme(
    text = element_text(family = "Helvetica", color = "black"),
    strip.background = element_blank(),
    axis.line = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 7)
  )

Fig2A

rm(hit_unfiltered_umap_layout, hit_unfiltered_umap_dt, D_jacc, nn_jacc)
```


# UMAP sensitivity analysis
```{r,fig.height = 3, fig.width = 7}
# Load umap run data generated in UMAP_sensitivity.R
umap_runs <- read.csv(file=paste0(intermediate_folder_path,"umap_runs.csv"))

# Visualize
Fig_S2_A <- ggplot(umap_runs, aes(V1, V2, color = factor(interval_label,levels = c("birth", "3mo", "6mo", "1yr","1.5yr", "2yr", "3yr","Adult")))) +
  facet_grid(min_dist ~ n_neighbors + metric) +
  theme_bw() +
  geom_point(size = 0.1,alpha=0.1) + 
  stat_ellipse(level = 0.95, linetype = "longdash")  + 
  scale_color_manual(name = "Age group", values = pal1[c(3,4,5,7,8,9,1,10)], guide = guide_legend(nrow=1)) +
  theme(text=element_text(family = "Helvetica",color = "black"),
        strip.background = element_blank(),
        axis.line = element_blank(), 
        axis.text = element_blank(),
        axis.ticks = element_blank(),panel.grid = element_blank(),
        legend.title = element_text(size=7),
        legend.text = element_text(size=7),
        strip.text = element_text(size=7),
        axis.title.x.top = element_text(size=7),
        axis.title.y.right = element_text(size=7),
        legend.position = "bottom") +
        xlab("n neighbors") +
        scale_y_continuous(position = "right") +
        scale_x_continuous(position = "top") +
        ylab("Minimum distance")
Fig_S2_A
```
# Quantitative analysis of UMAP parameters
```{r}
# silhouette score (global and by group) for every run
sil_summary <- umap_runs %>%
  group_by(n_neighbors, min_dist, metric, seed) %>%
  group_modify(~{
    d   <- dist(.x[, c("V1","V2")])
    grp <- factor(.x$interval_label)  
    si  <- cluster::silhouette(as.integer(grp), d)

    tibble(
      mean_sil = mean(si[, "sil_width"], na.rm = TRUE),
      n        = nrow(.x)
    )
  }) %>%
  ungroup()

# Per-group silhouettes 
sil_by_group <- umap_runs %>%
  group_by(n_neighbors, min_dist, metric, seed) %>%
  group_modify(~{
    d  <- dist(.x[, c("V1","V2")])
    grp <- factor(.x$interval_label) 
    si <- cluster::silhouette(as.integer(grp), d)
    s  <- as.data.frame(si)
    s$interval_label <- .x$interval_label
    s %>% group_by(interval_label) %>%
      summarise(group_sil = mean(sil_width), .groups = "drop")
  }) %>%
  ungroup()

# Summaries you can report (median across seeds)
sil_overview <- sil_summary %>%
  group_by(n_neighbors, min_dist, metric) %>%
  summarise(
    median_sil = median(mean_sil),
    IQR_sil    = IQR(mean_sil),
    .groups = "drop"
  )

#Procrustes stability across seeds (same parameters)
# Compare each seed to the smallest seed value as the "baseline" for a given parameter triple
procrustes_stability <- umap_runs %>%
  group_by(n_neighbors, min_dist, metric) %>%
  group_modify(~{
    g <- .x
    base_seed <- min(g$seed)
    base <- g %>% filter(seed == base_seed) %>% arrange(key)
    others <- g %>% filter(seed != base_seed)
    if (nrow(others) == 0) return(tibble(seed = integer(), t0 = numeric()))
    others %>%
      group_by(seed) %>%
      group_modify(~{
        h <- .x %>% arrange(key)
        # Procrustes test (no permutations): t0 is a correlation-like similarity [0,1]
        t0 <- vegan::protest(base[, c("V1","V2")], h[, c("V1","V2")], permutations = 0)$t0
        tibble(t0 = t0)
      }) %>%
      ungroup()
  }) %>%
  ungroup()

procrustes_overview <- procrustes_stability %>%
  group_by(n_neighbors, min_dist, metric) %>%
  summarise(
    median_t0 = median(t0),
    IQR_t0    = IQR(t0),
    .groups = "drop"
  )
```

# UMAP variations and Alternative analysis

# Fig S2B make UMAP using VirScores instead
```{r}
# Create long DF
adult_child_virscores<-AVARS_virscores %>% filter(identifier %in% IgG_Data$identifier) %>% left_join(IgG_Data %>% select(c(identifier,donor,time_interval,Age_Years, age_group, interval_label))) %>% mutate(age_group=ifelse(age_group==">1yr","1-3yr",age_group))%>% bind_rows(AVARS_virscores %>% filter(identifier %in% combined_metadata_dist$identifier[combined_metadata_dist$adult_cross==TRUE])%>% left_join(combined_metadata_dist %>% mutate(time_interval=40,age_group="adult",interval_label="Adult")%>% select(c(identifier,donor,time_interval,Age_Years, age_group, interval_label))))

# Add sample size to label
temp <- adult_child_virscores %>%
  dplyr::select(identifier, interval_label) %>%
  dplyr::distinct() %>%
  dplyr::group_by(interval_label) %>%
  dplyr::summarise(n_in_group = dplyr::n(), .groups = "drop") %>%
  dplyr::mutate(interval_label_updated = paste0(interval_label, " (", n_in_group, ")"))

adult_child_virscores <- adult_child_virscores %>%
  dplyr::left_join(temp %>% dplyr::select(-n_in_group), by = "interval_label")
rm(temp)

# Sample metadata (One row per identfier)
sample_meta <- adult_child_virscores %>% 
  dplyr::group_by(donor, identifier, Age_Years, time_interval, interval_label, interval_label_updated) %>% 
  dplyr::summarise(sum_virscore = sum(VirScore), .groups = "drop") %>%  
  dplyr::mutate(key = identifier)

# Build species-level matrix (rows = identifier)
vir_unfiltered_matrix <- adult_child_virscores %>%
  dplyr::mutate(key = identifier) %>% 
  tidyr::pivot_wider(id_cols = key, names_from = species, values_from = VirScore, values_fill = 0) %>% 
  tibble::column_to_rownames(var = "key") %>% 
  as.matrix()

# Drop uninformative columns (zero variance)
keep <- apply(vir_unfiltered_matrix, 2, sd, na.rm = TRUE) > 0
vir_unfiltered_matrix <- vir_unfiltered_matrix[, keep, drop = FALSE]
rm(keep)

# Drop very rare species (present in <5 samples)
prev <- colSums(vir_unfiltered_matrix != 0)
keep <- prev >= 5
vir_unfiltered_matrix <- vir_unfiltered_matrix[, keep, drop = FALSE]
rm(prev, keep)

# Row-normalize to emphasize profile composition (recommended with cosine)
rs <- rowSums(vir_unfiltered_matrix)
vir_unfiltered_matrix <- sweep(vir_unfiltered_matrix, 1, rs, "/")
vir_unfiltered_matrix[is.na(vir_unfiltered_matrix)] <- 0

# UMAP with uwot (cosine)
set.seed(1)
vir_unfiltered_layout <- uwot::umap(
  vir_unfiltered_matrix,
  n_neighbors = 15,
  n_components = 2,
  metric = "cosine",
  n_epochs = 200,
  init = "spectral",
  min_dist = 0.1,
  verbose = FALSE
)

# Format the UMAP layout for plotting
vir_unfiltered_umap_dt <- as.data.frame(vir_unfiltered_layout) %>%
  tibble::rownames_to_column(var = "key") %>%
  dplyr::left_join(sample_meta, by = "key")

Fig_S2_B <- ggplot(
  vir_unfiltered_umap_dt,
  aes(
    x = V1, y = V2,
    color = factor(
      interval_label_updated,
      levels = c(
        "birth (25)", "3mo (22)", "6mo (23)", "1yr (25)",
        "1.5yr (22)", "2yr (20)", "3yr (23)", "Adult (559)"
      )
    )
  )
) +
  geom_point(size = 0.5, alpha = 0.7) +
  stat_ellipse(level = 0.95, linetype = "longdash", type = "t") +
  scale_color_manual(name = "Age group", values = pal1[c(3,4,5,7,8,9,1,10)], guide = guide_legend(nrow = 2)) +
  theme_bw() +
  theme(
    text = element_text(family = "Helvetica", color = "black"),
    strip.background = element_blank(),
    axis.line = element_blank(),
    axis.title = element_text(size = 7),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 7),
    legend.position = "none"
  ) +
  labs(x = "V1", y = "V2", color = "Age group")

Fig_S2_B

rm(vir_unfiltered_layout, vir_unfiltered_umap_dt)

```




```{r}
# Class-imbalance sensitivity: downsample each age group to the smallest group
set.seed(5)
size_per_group <- min(table(sample_meta$interval_label))

balanced_keys <- sample_meta %>%
  dplyr::group_by(interval_label) %>%
  dplyr::slice_sample(n = size_per_group) %>%
  dplyr::pull(key)

X_bal <- hit_unfiltered_matrix[balanced_keys, , drop = FALSE]
smeta_bal <- sample_meta %>% dplyr::filter(key %in% balanced_keys)

# Helper: build exact kNN from a distance matrix (Jaccard workaround for uwot)
knn_from_dist <- function(D, k) {
  n <- nrow(D)
  k <- min(k, n - 1)

  idx <- t(vapply(seq_len(n), function(i) {
    ord <- order(D[i, ], na.last = NA)
    ord <- ord[ord != i]
    ord[seq_len(k)]
  }, integer(k)))

  dist <- t(vapply(seq_len(n), function(i) {
    D[i, idx[i, ]]
  }, numeric(k)))

  list(idx = idx, dist = dist)
}

# Run one representative UMAP on the balanced set (consistent with main fig: uwot + Jaccard neighbors)
set.seed(1)
D_jacc_bal <- as.matrix(vegan::vegdist(X_bal, method = "jaccard", binary = TRUE))
storage.mode(D_jacc_bal) <- "double"
nn_jacc_bal <- knn_from_dist(D_jacc_bal, k = 15)

umap_bal_layout <- uwot::umap(
  X = X_bal,
  n_neighbors = 15,
  n_components = 2,
  nn_method = nn_jacc_bal,   # <- Jaccard-based neighbors
  n_epochs = 200,
  init = "spectral",
  min_dist = 0.1,
  verbose = FALSE
)

umap_bal <- as.data.frame(umap_bal_layout) %>%
  tibble::rownames_to_column(var = "key") %>%
  dplyr::left_join(smeta_bal %>% dplyr::select(key, interval_label), by = "key")

Fig_S2_C <- ggplot(
  umap_bal,
  aes(
    V1, V2,
    color = factor(interval_label, levels = c("birth", "3mo", "6mo", "1yr","1.5yr", "2yr", "3yr","Adult"))
  )
) +
  geom_point(size = 0.5, alpha = 0.7) +
  stat_ellipse(level = 0.95, linetype = "longdash", type = "t") +
  scale_color_manual(name = "Age group", values = pal1[c(3,4,5,7,8,9,1,10)], guide = guide_legend(ncol = 2)) +
  theme_bw() +
  theme(
    text = element_text(family = "Helvetica", color = "black"),
    strip.background = element_blank(),
    axis.line = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 7),
    strip.text = element_text(size = 7),
    axis.title.x = element_text(size = 7),
    axis.title.y = element_text(size = 7),
    legend.position = "bottom"
  ) +
  labs(x = "V1", y = "V2", color = "Age group")

Fig_S2_C

rm(D_jacc_bal, nn_jacc_bal, umap_bal_layout)
```
# PERMANOVA on Jaccard distances with pemutations stratified by donor. Data frames generated in "High_Dimensional_Validation.R"
```{r}
# Align sample order
sample_meta <- sample_meta %>% slice(match(rownames(hit_unfiltered_matrix), key))
stopifnot(all(sample_meta$key == rownames(hit_unfiltered_matrix)))

D_jacc <- read.csv(file=paste0(intermediate_folder_path,"Jaccard_Distance_Matrix.csv"))
adon_age <- read.csv(paste0(intermediate_folder_path,"PERMANOVA_donor_stratified.csv"))
adon_age # First row of output is effect of interval_label, second is residual, and third is total

# Dispersion (are groups just more/less spread out?)
load(file=paste0(intermediate_folder_path,"betadisper.rdata"))
anova(bd)            # overall differences in dispersion
permutest(bd, permutations = 999)
TukeyHSD(bd)
```

# Fig S2D
```{r}
# Label-permutation null (permute age labels, keep distances fixed)
set.seed(10)
n_perm <- 500
perm_R2 <- read.csv(paste0(intermediate_folder_path,"label_permutation_null_R2.csv"))
obs_R2 <- adon_age$R2[1]
perm_p <- mean(perm_R2 >= obs_R2)
perm_p

Fig_S2_D <- ggplot(perm_R2) + geom_histogram(aes(x=x),bins = 500,fill=pal1[1])  +geom_vline(data=NULL,aes(xintercept = obs_R2),linetype="dotted") + 
  theme_bw() +
  labs(x=expression(paste("Effect size (", R^{2},")"))) +
  theme(text=element_text(family = "Helvetica",color = "black"),
        strip.background = element_blank(),
        axis.line = element_blank(), 
        axis.text = element_text(size=7),
        panel.grid = element_blank(),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7))

Fig_S2_D
```

# Reviewer requested "Is there evidence of kids aged 10-15 gaining profiles similar to adults? Does the PCA plot from Fig 2 look different once you include older children? When do kids focus on the common epitopes seen in adults?"
```{r}
# Look at other pediatric cohorts at discrete times
temp_identifiers <- combined_metadata_dist %>%
  dplyr::filter(ped_long_2 == TRUE) %>%
  dplyr::mutate(age_rounded = round(Age_Years, 0)) %>%
  dplyr::filter(Age_Years <= 0.85 | age_rounded %in% c(5, 8) | Age_Years > 10) %>%
  dplyr::mutate(age_rounded = ifelse(Age_Years > 10, 14, age_rounded)) %>%
  dplyr::mutate(age_rounded = ifelse(Age_Years <= 0.85, 0.5, age_rounded)) %>%
  dplyr::mutate(dist_from_age_round = abs(Age_Years - age_rounded)) %>%
  dplyr::group_by(donor, age_rounded) %>%
  dplyr::arrange(dist_from_age_round) %>%
  dplyr::slice_head(n = 1) %>%
  dplyr::mutate(interval_label = ifelse(age_rounded == 14, "10-14yrs", paste0(age_rounded, "yr"))) %>%
  dplyr::mutate(interval_label = ifelse(age_rounded == 0.5, "<1yr", interval_label)) %>%
  dplyr::ungroup()

# Create long DF
adult_older_child_longzs <- all_hits_combined %>%
  dplyr::select(c(id, all_of(temp_identifiers$identifier))) %>%
  tidyr::pivot_longer(-id, names_to = "identifier", values_to = "hit") %>%
  dplyr::left_join(temp_identifiers %>% dplyr::select(c(identifier, donor, Age_Years, interval_label)), by = "identifier") %>%
  dplyr::bind_rows(
    subset_frame_fxn(all_hits_combined, combined_metadata_dist, "adult_cross") %>%
      tidyr::pivot_longer(-id, names_to = "identifier", values_to = "hit") %>%
      dplyr::left_join(
        combined_metadata_dist %>%
          dplyr::mutate(interval_label = "Adult") %>%
          dplyr::select(c(identifier, donor, Age_Years, interval_label)),
        by = "identifier"
      )
  )

# Add sample size to label (clean count)
temp <- adult_older_child_longzs %>%
  dplyr::select(identifier, interval_label) %>%
  dplyr::distinct() %>%
  dplyr::group_by(interval_label) %>%
  dplyr::summarise(n_in_group = dplyr::n(), .groups = "drop") %>%
  dplyr::mutate(interval_label_updated = paste0(interval_label, " (", n_in_group, ")"))

adult_older_child_longzs <- adult_older_child_longzs %>%
  dplyr::left_join(temp %>% dplyr::select(-n_in_group), by = "interval_label")
rm(temp)

# Sample metadata 
sample_meta <- adult_older_child_longzs %>%
  dplyr::group_by(donor, identifier, Age_Years, interval_label, interval_label_updated) %>%
  dplyr::summarise(sum_hit = sum(hit), .groups = "drop") %>%
  dplyr::mutate(key = identifier)

# Build sample-by-peptide binary matrix (rows = identifier)
hit_unfiltered_matrix <- adult_older_child_longzs %>%
  dplyr::mutate(key = identifier) %>%
  tidyr::pivot_wider(id_cols = key, names_from = id, values_from = hit, values_fill = 0) %>%
  tibble::column_to_rownames(var = "key") %>%
  as.matrix()

# Drop peptides with extreme prevalence (>=5 and <= N-5)
cs <- colSums(hit_unfiltered_matrix)
keep <- cs >= 5 & cs <= (nrow(hit_unfiltered_matrix) - 5)
hit_unfiltered_matrix <- hit_unfiltered_matrix[, keep, drop = FALSE]
rm(cs, keep)

# Jaccard UMAP via uwot 
knn_from_dist <- function(D, k) {
  n <- nrow(D)
  k <- min(k, n - 1)

  idx <- t(vapply(seq_len(n), function(i) {
    ord <- order(D[i, ], na.last = NA)
    ord <- ord[ord != i]
    ord[seq_len(k)]
  }, integer(k)))

  dist <- t(vapply(seq_len(n), function(i) {
    D[i, idx[i, ]]
  }, numeric(k)))

  list(idx = idx, dist = dist)
}

set.seed(10)
D_jacc <- as.matrix(vegan::vegdist(hit_unfiltered_matrix, method = "jaccard", binary = TRUE))
storage.mode(D_jacc) <- "double"
nn_jacc <- knn_from_dist(D_jacc, k = 15)

hit_unfiltered_umap_layout <- uwot::umap(
  X = hit_unfiltered_matrix,
  n_neighbors = 15,
  n_components = 2,
  nn_method = nn_jacc,
  n_epochs = 200,
  init = "spectral",
  min_dist = 0.1,
  verbose = FALSE
)

# Format the UMAP layout
hit_unfiltered_umap_dt <- as.data.frame(hit_unfiltered_umap_layout) %>%
  tibble::rownames_to_column(var = "key") %>%
  dplyr::left_join(sample_meta, by = "key")

ggplot(
  hit_unfiltered_umap_dt,
  aes(
    x = V1, y = V2,
    color = factor(
      interval_label_updated,
      levels = c("<1yr (24)", "5yr (19)", "8yr (12)", "10-14yrs (12)", "Adult (559)")
    )
  )
) +
  geom_point(size = 0.5, alpha = 0.5) +
  stat_ellipse(level = 0.95, linetype = "longdash", type = "t") +
  scale_color_manual(name = "Age group", values = c("#198861", "#7A5AC4", "#CC2B8D", "#C66525", pal1[10])) +
  theme_bw() +
  theme(
    text = element_text(family = "Helvetica", color = "black"),
    strip.background = element_blank(),
    axis.line = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 7)
  )

rm(adult_older_child_longzs, sample_meta, hit_unfiltered_matrix, hit_unfiltered_umap_layout, hit_unfiltered_umap_dt, D_jacc, nn_jacc)
```

# Calculate whether breadth is different in children and adults
# Overall
```{r}
# 1) Identify the sample identifiers included in each cross-sectional set
adult_ids <- subset_frame_fxn(all_zs_combined, combined_metadata_dist, "adult_cross") %>%
  select(-id) %>%
  colnames()

ped_ids_all <- subset_frame_fxn(all_zs_combined, combined_metadata_dist, "ped_cross") %>%
  select(-id) %>%
  colnames()

ped_ids_25_35 <- combined_metadata_dist %>%
  filter(identifier %in% ped_ids_all, Age_Years >= 2.5, Age_Years <= 3.5) %>%
  pull(identifier) %>%
  unique()

# Keep only identifiers that exist in the hit matrix
hit_sample_cols <- setdiff(colnames(all_hits_combined), "id")
adult_ids   <- intersect(adult_ids, hit_sample_cols)
ped_ids_25_35 <- intersect(ped_ids_25_35, hit_sample_cols)

# Compute total peptides targeted per sample from all_hits_combined
overall_breadth_df <- all_hits_combined %>%
  select(id, all_of(c(adult_ids, ped_ids_25_35))) %>%
  pivot_longer(
    cols = -id,
    names_to = "identifier",
    values_to = "hit"
  ) %>%
  mutate(
    hit = as.numeric(hit) %>% tidyr::replace_na(0)
  ) %>%
  group_by(identifier) %>%
  summarize(
    Epitopes_Rec = sum(hit > 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    age_group = case_when(
      identifier %in% adult_ids ~ "Adult",
      identifier %in% ped_ids_25_35 ~ "Child",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(age_group))

# sanity checks (should be ~559 adults and 27 children )
cat("n_adults =", sum(overall_breadth_df$age_group == "Adult"),
    "n_children_2.5_3.5 =", sum(overall_breadth_df$age_group == "Child"), "\n")

# Two-sided Wilcoxon rank-sum test
res_overall <- wilcox.test(
  Epitopes_Rec ~ age_group,
  data = overall_breadth_df,
  alternative = "two.sided",
  exact = FALSE
)

res_overall
cat("W =", unname(res_overall$statistic), "p =", res_overall$p.value, "\n")
```

# Using VirScores
# Prepare data for composition (2B)
```{r}
# Create data frame for cross sectional data sets that defines exposed v not exposed
temp_z<-subset_frame_fxn(all_zs_combined,combined_metadata_dist,"adult_cross")
temp_z<-temp_z %>% left_join(Viral_Annotations %>% select(c(id,public_epitope_2015,species,family)))
temp_z<-temp_z %>% pivot_longer(-c(id,public_epitope_2015,species,family),names_to = "identifier",values_to = "ZScore")
# group by person and virus to get exposure variables
adult_exposure_vars_virscore <- temp_z %>% filter(ZScore>=3.5) %>% mutate(public_epitope_hit=ifelse(public_epitope_2015==TRUE,1,0)) %>% mutate(Z_above_5=ifelse(ZScore>=5,1,0)) %>% group_by(species,family,identifier) %>% summarise(n_public_hits=sum(public_epitope_hit),n_z_above_5=sum(Z_above_5),n_tot_hits=max(row_number()))
adult_exposure_vars_virscore <- adult_exposure_vars_virscore %>% left_join(AVARS_virscores)
n_adults<-length(unique(adult_exposure_vars_virscore$identifier)) # 559 adults
# how many exposed
temp<-Viral_Annotations %>% group_by(species,family) %>% summarise(n_lib_peptides=max(row_number()))
adult_exposed_summary_virscore<-adult_exposure_vars_virscore %>% mutate(exposed=ifelse((n_public_hits>=1 & n_z_above_5>=1) & VirScore>=2,1,0)) %>% filter(exposed==1) %>% group_by(species,family)%>% summarise(n_exposed=sum(exposed),avg_pub_hits=mean(n_public_hits),avg_z_over_5=mean(n_z_above_5),mean_hits=mean(n_tot_hits),mean_virscore=mean(VirScore)) %>% mutate(n_partic=n_adults) %>% left_join(temp)
# breadth of response among exposed

# Ped
temp_z<-subset_frame_fxn(all_zs_combined,combined_metadata_dist,"ped_cross")
temp_z<-temp_z %>% left_join(Viral_Annotations %>% select(c(id,public_epitope_2015,species,family)))
temp_z<-temp_z %>% pivot_longer(-c(id,public_epitope_2015,species,family),names_to = "identifier",values_to = "ZScore")
# group by person and virus to get exposure variables
# group by person and virus to get exposure variables
ped_exposure_vars_virscore <- temp_z %>% filter(ZScore>=3.5) %>% mutate(public_epitope_hit=ifelse(public_epitope_2015==TRUE,1,0)) %>% mutate(Z_above_5=ifelse(ZScore>=5,1,0)) %>% group_by(species,family,identifier) %>% summarise(n_public_hits=sum(public_epitope_hit),n_z_above_5=sum(Z_above_5),n_tot_hits=max(row_number()))
ped_exposure_vars_virscore <- ped_exposure_vars_virscore %>% left_join(AVARS_virscores)
n_ped<-length(unique(ped_exposure_vars_virscore$identifier)) # 27 peds
# how many exposed
temp<-Viral_Annotations %>% group_by(species,family) %>% summarise(n_lib_peptides=max(row_number()))
ped_exposed_summary_virscore<-ped_exposure_vars_virscore %>% mutate(exposed=ifelse((n_public_hits>=1 & n_z_above_5>=1) & VirScore>=2,1,0)) %>% filter(exposed==1) %>% group_by(species,family)%>% summarise(n_exposed=sum(exposed),avg_pub_hits=mean(n_public_hits),avg_z_over_5=mean(n_z_above_5),mean_hits=mean(n_tot_hits),mean_virscore=mean(VirScore)) %>% mutate(n_partic=n_ped) %>% left_join(temp)
rm(temp,temp_z)
length(unique(ped_exposure_vars_virscore$identifier))
length(unique(adult_exposure_vars_virscore$identifier))
```

```{r}
exposure_vars_combined_virscore<-bind_rows(ped_exposure_vars_virscore,adult_exposure_vars_virscore) %>% mutate(exposed=ifelse((n_public_hits>=1 & n_z_above_5>=1) & VirScore>=2,1,0)) %>% filter(exposed==1) %>% left_join(combined_metadata_dist %>% select(c(identifier,age_group,Age_Years,Sex)))
exposure_vars_combined_virscore
```

# Average proportions in children and adults
```{r,fig.width=11,fig.height=6}
family_prop<- bind_rows(family_breakdown_prop_fn(subset_frame_fxn(all_hits_combined,combined_metadata_dist,"adult_cross"),Viral_Annotations) %>% left_join(combined_metadata_dist %>% select(c(identifier,age_group,Age_Years))), family_breakdown_prop_fn(subset_frame_fxn(all_hits_combined,combined_metadata_dist,"ped_cross"),Viral_Annotations) %>% left_join(combined_metadata_dist %>% select(c(identifier,age_group,Age_Years))))
family_prop<-family_prop %>% select(-tot_hits) %>% pivot_longer(-c(identifier,Age_Years,age_group),names_to = "family_label",values_to = "prop_of_rep")

# test for normality for each virus and age group
prop_normality_results<-tidyr::crossing("family_label"=unique(family_prop$family_label),"age_group"=unique(exposure_vars_combined_virscore$age_group))
prop_normality_results$shapirowilk_p=100
for (i in seq_along(prop_normality_results$age_group)) {
  tempdat<-family_prop %>%filter(age_group==prop_normality_results$age_group[i],family_label==prop_normality_results$family_label[i])
  prop_normality_results$shapirowilk_p[i] <-shapiro.test(tempdat$prop_of_rep)[[2]]
}
prop_normality_results %>% filter(shapirowilk_p<=0.05)
# none of the adult values are normal; use wilcox
wilcox_props<-data.frame("family_label"=unique(family_prop$family_label))
for (i in seq_along(wilcox_props$family_label)) {
  wilcox_props$wilk_p[i]<-wilcox.test(prop_of_rep~age_group,data=family_prop %>% filter(family_label==wilcox_props$family_label[i]))[[3]]
}

wilcox_props$adjust.wilcox_p=p.adjust(wilcox_props$wilk_p, method="BH")
wilcox_props %>% filter(adjust.wilcox_p<0.05)
wilcox_props<-wilcox_props%>% mutate(signif=ifelse(adjust.wilcox_p<0.05,"*",""))%>% mutate(signif=ifelse(adjust.wilcox_p<0.01,"**",signif))%>% mutate(signif=ifelse(adjust.wilcox_p<0.001,"***",signif))

family_prop<-family_prop %>% left_join(wilcox_props %>% select(c(family_label,signif)))%>% mutate(family_label_2=paste0(family_label,signif))
family_prop
```

```{r}
exposure_normality_results_virscore<-tidyr::crossing("species"=unique(exposure_vars_combined_virscore$species),"age_group"=unique(exposure_vars_combined_virscore$age_group))
exposure_normality_results_virscore$shapirowilkp=100
exposure_normality_results_virscore$n_exposed=100
for (i in seq_along(exposure_normality_results_virscore$age_group)) {
  tempdat<-exposure_vars_combined_virscore %>%filter(age_group==exposure_normality_results_virscore$age_group[i],species==exposure_normality_results_virscore$species[i])
  exposure_normality_results_virscore$n_exposed[i]=nrow(tempdat)
  if(nrow(tempdat)<3){
    exposure_normality_results_virscore$shapirowilkp[i]=NA
  } else {
    exposure_normality_results_virscore$shapirowilkp[i] <-shapiro.test(tempdat$VirScore)[[2]]
  }
}
head(exposure_normality_results_virscore)
exposure_normality_results_virscore %>% filter(shapirowilkp>0.05)
rm(tempdat)
# some pediatric responses are normally distributed. Few adult are so use wilcoxon rank sum test


```

# Identify viruses with at least 5 exposures per group and perform  wilcoxon test
```{r}
temp_viruses<-exposure_normality_results_virscore %>% filter(n_exposed>=5) %>% group_by(species) %>% summarise(tempn=max(row_number())) %>% filter(tempn==2) %>% select(species) %>% unlist() %>% unname()
M_W_Ep_test_virscore<-data.frame("species"=c(temp_viruses),"wilcox_p"=100)
for (i in seq_along(M_W_Ep_test_virscore$species)) {
  tempdat<-exposure_vars_combined_virscore %>%filter(species==M_W_Ep_test_virscore$species[i])
  M_W_Ep_test_virscore$wilcox_p[i] <-wilcox.test(VirScore~age_group,data=tempdat)[[3]]
}
M_W_Ep_test_virscore$adjust.wilcox_p=p.adjust(M_W_Ep_test_virscore$wilcox_p, method="BH")
M_W_Ep_test_virscore %>% filter(adjust.wilcox_p>0.05)
M_W_Ep_test_virscore%>% filter(adjust.wilcox_p<=0.05)

```

```{r}
#arrange by difference
temp_ord<-M_W_Ep_test_virscore %>% filter(adjust.wilcox_p<0.05) %>% left_join(exposure_vars_combined_virscore) %>% group_by(species,wilcox_p,adjust.wilcox_p,family,age_group) %>% summarise(mean_virscore=mean(VirScore)) %>% pivot_wider(id_cols = c(species,wilcox_p,adjust.wilcox_p,family),names_from = age_group,values_from = mean_virscore,names_prefix = "mean_virscore_") %>% mutate(age_dif=mean_virscore_Adult-mean_virscore_Pediatric) %>% arrange(age_dif) %>% select(c(species,age_dif))
plot_mean_species<-M_W_Ep_test_virscore %>% filter(adjust.wilcox_p<0.05) %>% left_join(exposure_vars_combined_virscore) %>% group_by(species,wilcox_p,adjust.wilcox_p,family,age_group) %>% summarise(mean_virscore=mean(VirScore),sd_virscore=sd(VirScore),n_in_group=max(row_number())) %>% mutate(sem=sd_virscore/sqrt(n_in_group),t_crit_value=qt(.025,n_in_group-1)*(-1))

# Output sample sizes for figure 2c legend
fig2c_n_for_legend <- plot_mean_species %>%
  dplyr::select(species, age_group, n_in_group) %>%
  tidyr::pivot_wider(
    names_from = age_group,
    values_from = n_in_group
  ) %>%
  dplyr::mutate(
    legend_text = paste0(species, ", adult n = ", Adult, ", pediatric n = ", Pediatric)
  ) %>%
  dplyr::arrange(factor(species, levels = temp_ord$species))

cat(paste(fig2c_n_for_legend$legend_text, collapse = "; "))
```





```{r,fig.width=7}
Fig2B<-ggplot(family_prop)+geom_density(alpha=0.6,aes(x=prop_of_rep*100,y=..scaled..,fill=age_group,color=age_group))+facet_wrap(~factor(family_label_2,levels=c("Herpesviridae***", "Orthomyxoviridae**", "Pneumoviridae**","Coronaviridae*","Paramyxoviridae**", "Picornaviridae**","Adenoviridae**", "Other***")),nrow=2)+scale_fill_manual(name="",values = pal1[c(10,6)])+scale_color_manual(name="",values = pal1[c(10,6)])+theme_minimal()+theme(text = element_text(family = "Helvetica", size=7),panel.grid = element_blank(),legend.text  = element_text(size=7),panel.background = element_blank(),panel.border = element_rect(fill=NA,colour = "black"),strip.text.x = element_text(size=7),axis.text = element_text(size=7),axis.title = element_text(size=7))+ylab("Scaled density")+xlab("Percent of antiviral repertoire")
Fig2B
```

```{r}
Fig2C<-ggplot(plot_mean_species)+geom_pointrange(alpha=0.6,position=ggstance::position_dodgev(height=0.3),pch=20, aes(x=mean_virscore,y=factor(species,levels=c(temp_ord$species)),xmin=mean_virscore-t_crit_value*sem,xmax=mean_virscore+t_crit_value*sem,color=age_group))+scale_color_manual(name="Age group",values = pal1[c(10,6)])+theme_minimal()+theme(text=element_text(family="Helvetica"),axis.title.x = element_text(size=7),axis.text=element_text(size=7),legend.position = "none",panel.grid.major.x = element_blank(),panel.grid.minor.x = element_blank(),axis.title.y = element_blank(),plot.margin=unit(c(1,0,0,0), 'cm'))+xlab("Mean Virscore")
Fig2C
```

```{r,fig.height=7,fig.width=6}
ggarrange(ggarrange(ggplot(NULL)+theme_void(), Fig2A, ggplot(NULL)+theme_void(), widths = c(0.4,8,0.4),nrow = 1), Fig2B,Fig2C+theme(plot.margin = unit(c(0, 0.5, 0.5, 0.5), "cm")),labels=c("a","b","c"),heights=c(1.75,2,2),ncol=1,font.label = list(family="Helvetica",size=8))
ggsave(file=paste0(folder_path_save,"/AVARS_Figure_2_Final.pdf"))
dev.off()
```
# Does Panel C look similar with downsampling
```{r}
temp_size<-length(combined_metadata_dist$identifier[combined_metadata_dist$ped_cross==TRUE]) # downsample adults to the number of pediatric donors
set.seed(5) 
adults_downsampled <- combined_metadata_dist %>% filter(adult_cross==TRUE) %>% sample_n(temp_size) %>% select(identifier) %>% unlist() %>% unname()


family_prop_downsampled <- bind_rows(family_breakdown_prop_fn(all_hits_combined %>% select(c(id,all_of(adults_downsampled))),Viral_Annotations) %>% left_join(combined_metadata_dist %>% select(c(identifier,age_group,Age_Years))), family_breakdown_prop_fn(subset_frame_fxn(all_hits_combined,combined_metadata_dist,"ped_cross"),Viral_Annotations) %>% left_join(combined_metadata_dist %>% select(c(identifier,age_group,Age_Years))))
family_prop_downsampled<-family_prop_downsampled %>% select(-tot_hits) %>% pivot_longer(-c(identifier,Age_Years,age_group),names_to = "family_label",values_to = "prop_of_rep")

# test for normality for each virus and age group
prop_normality_results_downsampled<-tidyr::crossing("family_label"=unique(family_prop_downsampled$family_label),"age_group"=unique(exposure_vars_combined_virscore$age_group))
prop_normality_results_downsampled$shapirowilk_p=100
for (i in seq_along(prop_normality_results_downsampled$age_group)) {
  tempdat<-family_prop_downsampled %>%filter(age_group==prop_normality_results_downsampled$age_group[i],family_label==prop_normality_results_downsampled$family_label[i])
  prop_normality_results_downsampled$shapirowilk_p[i] <-shapiro.test(tempdat$prop_of_rep)[[2]]
}
prop_normality_results_downsampled %>% filter(shapirowilk_p<=0.05)
# Not normal; use wilcox
wilcox_props_downsampled<-data.frame("family_label"=unique(family_prop_downsampled$family_label))
for (i in seq_along(wilcox_props_downsampled$family_label)) {
  wilcox_props_downsampled$wilk_p[i]<-wilcox.test(prop_of_rep~age_group,data=family_prop_downsampled %>% filter(family_label==wilcox_props_downsampled$family_label[i]))[[3]]
}

wilcox_props_downsampled$adjust.wilcox_p=p.adjust(wilcox_props_downsampled$wilk_p, method="BH")
wilcox_props_downsampled %>% filter(adjust.wilcox_p<0.05)
wilcox_props_downsampled<-wilcox_props_downsampled%>% mutate(signif=ifelse(wilk_p<0.05,"*",""))%>% mutate(signif=ifelse(wilk_p<0.01,"**",signif))%>% mutate(signif=ifelse(wilk_p<0.001,"***",signif))

family_prop_downsampled<-family_prop_downsampled %>% left_join(wilcox_props_downsampled %>% select(c(family_label,signif)))%>% mutate(family_label_2=paste0(family_label,signif))

family_prop_downsampled

Fig_S2_E <- ggplot(family_prop_downsampled)+geom_density(alpha=0.6,aes(x=prop_of_rep*100,y=..scaled..,fill=age_group,color=age_group))+facet_wrap(~factor(family_label_2,levels=c("Herpesviridae***", "Orthomyxoviridae*", "Pneumoviridae","Coronaviridae","Paramyxoviridae*", "Picornaviridae","Adenoviridae", "Other***")),nrow=2)+scale_fill_manual(name="",values = pal1[c(10,6)])+scale_color_manual(name="",values = pal1[c(10,6)])+theme_minimal()+theme(text = element_text(family = "Helvetica", size=7),panel.grid = element_blank(),legend.text  = element_text(size=7),panel.background = element_blank(),panel.border = element_rect(fill=NA,colour = "black"),strip.text.x = element_text(size=7),axis.text = element_text(size=7),axis.title = element_text(size=7))+ylab("Scaled density")+xlab("Percent of antiviral repertoire")
Fig_S2_E
```
# Patterns are similar with downsampling; lose some statistical significance

# Do the same for panel D - this becomes Fig S2f
```{r}
set.seed(5)

# 1) Species eligible for testing: >=5 per age_group before downsampling
eligible_species <- exposure_vars_combined_virscore %>%
  dplyr::count(species, age_group, name = "n") %>%
  dplyr::filter(n >= 5) %>%
  dplyr::count(species, name = "n_groups") %>%
  dplyr::filter(n_groups == 2) %>%
  dplyr::pull(species)

dat_eligible <- exposure_vars_combined_virscore %>%
  dplyr::filter(species %in% eligible_species) %>%
  dplyr::filter(age_group %in% c("Adult", "Pediatric"))

# 2) Downsample within each species to balance Adult/Pediatric
dat_bal <- dat_eligible %>%
  dplyr::group_by(species) %>%
  dplyr::group_modify(~{
    g <- .x
    n_min <- min(table(g$age_group))
    g %>%
      dplyr::group_by(age_group) %>%
      dplyr::slice_sample(n = n_min) %>%
      dplyr::ungroup()
  }) %>%
  dplyr::ungroup()

# Store the balanced n used per age group for each species
species_n_bal <- dat_bal %>%
  dplyr::count(species, age_group, name = "n_in_group") %>%
  dplyr::group_by(species) %>%
  dplyr::summarise(
    n_balanced = unique(n_in_group)[1],
    .groups = "drop"
  )

# 3) Wilcoxon per species on the balanced data + BH correction across species
mw_bal <- dat_bal %>%
  dplyr::group_by(species) %>%
  dplyr::summarise(
    wilcox_p = wilcox.test(
      VirScore ~ age_group,
      data = dplyr::cur_data(),
      alternative = "two.sided",
      exact = FALSE
    )$p.value,
    .groups = "drop"
  ) %>%
  dplyr::mutate(adjust_wilcox_p = p.adjust(wilcox_p, method = "BH"))

sig_species_bal <- mw_bal %>%
  dplyr::filter(wilcox_p < 0.05) %>%
  dplyr::pull(species)

# If nothing survives the significance threshold, this will just plot nothing
dat_bal_sig <- dat_bal %>%
  dplyr::filter(species %in% sig_species_bal)

# 4) Summaries for plotting: mean +/- t * SEM
plot_mean_species_bal_sig <- dat_bal_sig %>%
  dplyr::group_by(species, family, age_group) %>%
  dplyr::summarise(
    mean_virscore = mean(VirScore, na.rm = TRUE),
    sd_virscore   = sd(VirScore, na.rm = TRUE),
    n_in_group    = dplyr::n(),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    sem = sd_virscore / sqrt(n_in_group),
    t_crit_value = qt(.025, n_in_group - 1) * (-1)
  ) %>%
  dplyr::left_join(species_n_bal, by = "species") %>%
  dplyr::mutate(
    species_label = paste0(species, " (n = ", n_balanced, ")")
  )

# 5) Order species by Adult - Pediatric mean difference
temp_ord_bal_sig <- plot_mean_species_bal_sig %>%
  dplyr::select(species, species_label, age_group, mean_virscore) %>%
  tidyr::pivot_wider(names_from = age_group, values_from = mean_virscore) %>%
  dplyr::mutate(age_dif = Adult - Pediatric) %>%
  dplyr::arrange(age_dif)

species_label_levels_bal_sig <- temp_ord_bal_sig$species_label

Fig_S2_F <- ggplot(plot_mean_species_bal_sig) +
  geom_pointrange(
    alpha = 0.6,
    position = ggstance::position_dodgev(height = 0.3),
    pch = 20,
    aes(
      x = mean_virscore,
      y = factor(species_label, levels = species_label_levels_bal_sig),
      xmin = mean_virscore - t_crit_value * sem,
      xmax = mean_virscore + t_crit_value * sem,
      color = age_group
    )
  ) +
  scale_color_manual(name = "Age group", values = pal1[c(10, 6)]) +
  theme_minimal() +
  theme(
    text = element_text(family = "Helvetica", size = 7),
    axis.title.x = element_text(size = 7),
    axis.text.y  = element_text(size = 7),
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    axis.title.y = element_blank(),
    plot.margin = unit(c(1, 0, 0, 0), "cm")
  ) +
  xlab("Mean VirScore")

Fig_S2_F

```




# Figure S2
# Sensitivity analysis and alternatives to UMAP

```{r,fig.height=6.75,fig.width=7}
ggarrange(Fig_S2_A, ggarrange(Fig_S2_B, Fig_S2_C + theme (legend.position = "none"),Fig_S2_D, nrow=1, font.label = list(family="Helvetica",size=8),labels = c("b","c","d")), ggarrange(Fig_S2_E + theme(axis.text = element_text(size=7),legend.position = "top"),Fig_S2_F,font.label = list(family="Helvetica",size=8),labels = c("e","f"),widths = c(2,1.5)), nrow = 3, heights = c(2,1.2,1.5),font.label = list(family="Helvetica",size=8),labels = c("a","",""))
ggsave(file=paste0(folder_path_save,"Supplemental Figures/AVARS_Figure_S2_Final.pdf"))
dev.off()
```


# Figure S1
## Panel A is UMAP generated for specific viral families
### Data preparation
```{r}
# Helper: build exact kNN from a distance matrix (Jaccard workaround for uwot)
knn_from_dist <- function(D, k) {
  n <- nrow(D)
  k <- min(k, n - 1)

  idx <- t(vapply(seq_len(n), function(i) {
    ord <- order(D[i, ], na.last = NA)
    ord <- ord[ord != i]        # drop self
    ord[seq_len(k)]
  }, integer(k)))

  dist <- t(vapply(seq_len(n), function(i) {
    D[i, idx[i, ]]
  }, numeric(k)))

  list(idx = idx, dist = dist)
}

# Helper: run family-specific UMAP (identifier key + prevalence filter + uwot + Jaccard via nn_method)
run_family_umap <- function(long_df, family_ids, Viral_Annotations, n_neighbors = 15) {

  # sample_meta (one row per identifier)
  sample_meta <- long_df %>%
    dplyr::filter(id %in% family_ids) %>%
    dplyr::group_by(donor, identifier, Age_Years, time_interval, interval_label, interval_label_updated) %>%
    dplyr::summarise(sum_hit = sum(hit), .groups = "drop") %>%
    dplyr::mutate(key = identifier)

  # peptide_meta (kept in case you still use it downstream)
  peptide_meta <- long_df %>%
    dplyr::filter(id %in% family_ids) %>%
    dplyr::left_join(Viral_Annotations %>% dplyr::select(id, family_label, species), by = "id") %>%
    dplyr::group_by(id, species, family_label) %>%
    dplyr::summarise(positive_sample = sum(hit), .groups = "drop")

  # build binary matrix (rows = identifier)
  X <- long_df %>%
    dplyr::filter(id %in% family_ids) %>%
    dplyr::mutate(key = identifier) %>%
    tidyr::pivot_wider(id_cols = key, names_from = id, values_from = hit, values_fill = 0) %>%
    tibble::column_to_rownames(var = "key") %>%
    as.matrix()

  # prevalence filter (consistent with Fig 2A)
  cs <- colSums(X)
  keep <- cs >= 5 & cs <= (nrow(X) - 5)
  X <- X[, keep, drop = FALSE]
  rm(cs, keep)

  # Jaccard distances + kNN
  D_jacc <- as.matrix(vegan::vegdist(X, method = "jaccard", binary = TRUE))
  storage.mode(D_jacc) <- "double"
  nn_jacc <- knn_from_dist(D_jacc, k = n_neighbors)

  # UMAP with uwot using precomputed neighbors
  set.seed(1)
  layout <- uwot::umap(
    X = X,
    n_neighbors = n_neighbors,
    n_components = 2,
    nn_method = nn_jacc,
    n_epochs = 200,
    init = "spectral",
    min_dist = 0.1,
    verbose = FALSE
  )

  umap_dt <- as.data.frame(layout) %>%
    tibble::rownames_to_column(var = "key") %>%
    dplyr::left_join(sample_meta, by = "key")

  list(
    umap_dt = umap_dt,
    sample_meta = sample_meta,
    peptide_meta = peptide_meta,
    X = X
  )
}

# Define family ID sets
herpes_ids <- Viral_Annotations$id[Viral_Annotations$family_label == "Herpesviridae"]
adeno_ids  <- Viral_Annotations$id[Viral_Annotations$family_label == "Adenoviridae"]
flu_ids    <- Viral_Annotations$id[Viral_Annotations$family_label == "Orthomyxoviridae"]
picorn_ids <- Viral_Annotations$id[Viral_Annotations$family_label == "Picornaviridae"]
cov_ids    <- Viral_Annotations$id[Viral_Annotations$family_label == "Coronaviridae"]
paramyx_ids<- Viral_Annotations$id[Viral_Annotations$family_label == "Paramyxoviridae"]

# Run UMAPs (each returns a list; we keep umap_dt for plotting)
hit_herpes_umap_dt  <- run_family_umap(adult_child_longzs, herpes_ids, Viral_Annotations)$umap_dt
hit_adeno_umap_dt   <- run_family_umap(adult_child_longzs, adeno_ids,  Viral_Annotations)$umap_dt
hit_flu_umap_dt     <- run_family_umap(adult_child_longzs, flu_ids,    Viral_Annotations)$umap_dt
hit_picorn_umap_dt  <- run_family_umap(adult_child_longzs, picorn_ids, Viral_Annotations)$umap_dt
hit_cov_umap_dt     <- run_family_umap(adult_child_longzs, cov_ids,    Viral_Annotations)$umap_dt
hit_paramyx_umap_dt <- run_family_umap(adult_child_longzs, paramyx_ids,Viral_Annotations)$umap_dt
```

### Make panel
```{r}
hit_umap_dt <- dplyr::bind_rows(
  list(
    "Herpesvirus"    = hit_herpes_umap_dt,
    "Adenovirus"     = hit_adeno_umap_dt,
    "Orthomyxovirus" = hit_flu_umap_dt,
    "Picornavirus"   = hit_picorn_umap_dt,
    "Coronavirus"    = hit_cov_umap_dt,
    "Paramyxovirus"  = hit_paramyx_umap_dt
  ),
  .id = "filter"
) %>%
  dplyr::mutate(
    filter = ordered(
      filter,
      levels = c("Herpesvirus", "Adenovirus","Orthomyxovirus","Picornavirus","Coronavirus","Paramyxovirus"),
      labels = c(
        paste0("Herpesviridae peptides\n(n = ", length(herpes_ids), ")"),
        paste0("Adenoviridae peptides\n(n = ", length(adeno_ids), ")"),
        paste0("Orthomyxoviridae peptides\n(n = ", length(flu_ids), ")"),
        paste0("Picornaviridae peptides\n(n = ", length(picorn_ids), ")"),
        paste0("Coronaviridae peptides\n(n = ", length(cov_ids), ")"),
        paste0("Paramyxoviridae peptides\n(n = ", length(paramyx_ids), ")")
      )
    )
  )

Fig_S1_A <- ggplot(
  hit_umap_dt,
  aes(
    x = V1, y = V2,
    color = factor(
      interval_label_updated,
      levels = c(
        "birth (25)", "3mo (22)", "6mo (23)", "1yr (25)",
        "1.5yr (22)", "2yr (20)", "3yr (23)", "Adult (559)"
      )
    )
  )
) +
  geom_point(size = 1.5, alpha = 0.5) +
  stat_ellipse(level = 0.95, linetype = "longdash", type = "t") +
  facet_wrap(facets = vars(filter), nrow = 2, scales = "free") +
  scale_color_manual(name = "Age group", values = c( pal1[c(3,4,5,7,8,9,1)], "grey65")) +
  theme_bw(base_size = 8) +
  theme(
    strip.background = element_blank(),
    axis.line = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    text = element_text(family = "Helvetica", colour = "black",size = 7),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 7),
    legend.position = "top"
  )

Fig_S1_A
```


## Combine and save
```{r,fig.width=7,fig.height=4}
Fig_S1_A
ggsave(file=paste0(folder_path_save,"/AVARS_Figure_S1_Final.pdf"))
dev.off()
```



