---
title: "R Notebook"
output: html_notebook
---

```{r}
easypackages::libraries(c("readr","readxl","dplyr","tidyr","ggplot2","ggpubr","UpSetR","cowplot","grid","ggplotify","patchwork","stringr","igraph","ggraph"))

# insert path to folders
directory_path <- "" # CHANGE THIS to path to director enclosing code and data sets
folder_path <- paste0(directory_path, "/Supplemental Tables/") # folder containing downloaded data sets
intermediate_folder_path <- paste0(directory_path, "/Intermediate Files Used in Plots/") # path containing intermediate outputs
folder_path_save <- paste0(directory_path, "/Complete Figures/")# define folder path to save images to
folder_path_code <- paste0(directory_path, "/Functions and General Analysis Code/") # change this to folder containing downloaded functions and code

# Load Supplemental Data
combined_metadata_dist <- read_csv(paste0(folder_path,"Data Table S1.csv"))
all_zs_combined <- read_csv(paste0(folder_path,"Data Table S2.csv"))
all_hits_combined <- read_csv(paste0(folder_path,"Data Table S3.csv"))
Viral_Annotations <- read_csv(paste0(folder_path,"Data Table S4.csv"))
IgG_Data <- read_csv(paste0(folder_path,"Data Table S6.csv"))
Vaccination_Data <- read_csv(paste0(folder_path,"Data Table S7.csv"))
Birth_Consec_Zs <- read_csv(paste0(intermediate_folder_path,"Birth_Consec_Zs.csv")) # Generated in "Generate Circle Plot Data.R"
age_at_exposure <- read_csv(paste0(intermediate_folder_path,"circle_plot_data_figure_5.csv")) # Generated in "Generate Circle Plot Data.R"
virus_key_5D <- read_csv(paste0(intermediate_folder_path,"virus_key_figure_5.csv")) # Generated in "Generate Circle Plot Data.R"


# Load custom functions
source(paste0(folder_path_code,"Immunoprint functions.R"))
cosine_sim <- function(x, y) {
  # robust cosine for numeric vectors
  num <- sum(x * y, na.rm = TRUE)
  den <- sqrt(sum(x * x, na.rm = TRUE)) * sqrt(sum(y * y, na.rm = TRUE))
  if (is.na(den) || den == 0) return(NA_real_)
  num / den
}

pal1<-c("#F87E2E", "#FFEC33", "#1FAA79", "#14C4F8", "#0077C8", "#9870F5", "#D53EE1", "#FF36B0", "#D72638", "#6E6E6E")
```

# Limit vaccination data to window sampled
```{r}
temp <- combined_metadata_dist %>% filter(donor %in% Vaccination_Data$donor) %>% group_by(donor) %>% summarise(max_age=max(Age_Years),min_age=min(Age_Years)) %>% filter(min_age==0)
Vaccination_Data <- Vaccination_Data %>% filter(donor %in% temp$donor)%>% left_join(temp) %>% filter(Age_Years<max_age)
Vaccination_Data
```
# Maternal Endogenous Labeling of Vaccine Antigen Responses
```{r}
vaccine_ids <- Viral_Annotations %>% filter((species=="Enterovirus C" & grepl("Poliovirus",Organism))|(species %in% c("Rotavirus A","Rotavirus B") & grepl("Human",Organism))|species %in% c("Measles morbillivirus","Rubivirus rubellae")|species %in% c("Influenza A virus","Influenza B virus")) %>% select(id) %>% unlist() %>% unname()

Mat_end_hits<-all_hits_combined %>% filter(id %in% vaccine_ids) %>%  select(c(id,all_of(IgG_Data$identifier))) %>% left_join(Viral_Annotations%>% select(c(id,species,Organism))) %>% pivot_longer(-c(id,species,Organism),names_to = "identifier",values_to = "hit") %>% left_join(IgG_Data %>% select(c(identifier,donor,time_interval,Age_Years, age_group, interval_label))) %>% left_join(all_zs_combined %>% filter(id %in% vaccine_ids) %>% select(c(id,all_of(IgG_Data$identifier))) %>% pivot_longer(-id,names_to = "identifier",values_to = "zscore")) %>% left_join(bound_Z(all_zs_combined %>% filter(id %in% vaccine_ids) %>% select(c(id,all_of(IgG_Data$identifier)))) %>% pivot_longer(-id,names_to = "identifier",values_to = "trunc_zscore"))

# select only individual + id for which there is a hit at birth
present_at_birth <- filter(Mat_end_hits, Age_Years == 0, hit == 1) %>%
  mutate(indiv_id_key = paste(donor, id)) %>%
  select(c(indiv_id_key))

# all possible trajectories
keep_maternal <- c("maternal_maternal_maternal_maternal", 
                   "maternal_missing_maternal_maternal")

rm_after_0 <- c("maternal_not maternal_maternal_maternal", 
                "maternal_not maternal_maternal_not maternal", 
                "maternal_not maternal_not maternal_maternal",
                "maternal_not maternal_not maternal_not maternal",
                "maternal_not maternal_missing_maternal",
                "maternal_not maternal_missing_not maternal",
                "maternal_missing_not maternal_maternal",
                "maternal_missing_not maternal_not maternal")

rm_after_1 <- c("maternal_maternal_not maternal_maternal", 
                "maternal_maternal_not maternal_not maternal",
                "maternal_maternal_missing_not maternal",
                "maternal_maternal_missing_maternal")

rm_after_2 <- c("maternal_maternal_maternal_not maternal", 
                "maternal_missing_maternal_not maternal")

# label epitopes
maternal_hits <- Mat_end_hits %>%
  filter(paste(donor, id) %in% present_at_birth$indiv_id_key, 
           # select only those epitopes present at birth
         time_interval <= 1) %>%  # select only timepoints less than 1yr
  mutate(key = paste(donor, time_interval, id)) %>%
  group_by(donor, id) %>%
  arrange(Age_Years) %>%
  mutate(delta_zscore = c(0, diff(zscore)), 
           # get difference from timepoint to timepoint to see whether z score is decreasing or increasing
         maternal = case_when(
           Age_Years == 0 ~ "maternal",
           hit == 1 & delta_zscore <= 0 ~ "maternal", 
           hit == 0 ~ "not maternal",
           hit == 1 & delta_zscore > 0 ~ "not maternal"
           # maternal antibodies: enriched at birth and declining afterwards
         )) %>%  
  select(-c(Age_Years, zscore, age_group, trunc_zscore, hit, delta_zscore, key, interval_label,identifier)) %>%
  pivot_wider(names_from = time_interval, values_from = maternal, values_fill = "missing") %>%
  mutate(trajectory = paste(`0`, `0.25`, `0.5`, `1`, sep = "_"),
         rm = case_when(
           trajectory %in% keep_maternal ~ "keep_maternal",
           trajectory %in% rm_after_0 ~ "rm_after_0",
           trajectory %in% rm_after_1 ~ "rm_after_1",
           trajectory %in% rm_after_2 ~ "rm_after_2")) %>%
  pivot_longer(cols = c(`0`, `0.25`, `0.5`, `1`), names_to = "time_interval", values_to = "maternal") %>%
  mutate(maternal = case_when(
    time_interval == 0 ~ "maternal",
    rm == "keep_maternal" ~ "maternal",
    rm == "rm_after_0" & time_interval > 0 ~ "not maternal",
    rm == "rm_after_1" & time_interval > 0.25 ~ "not maternal",
    rm == "rm_after_1" & time_interval <= 0.25 ~ "maternal",
    rm == "rm_after_2" & time_interval > 0.5 ~ "not maternal",
    rm == "rm_after_2" & time_interval <= 0.5 ~ "maternal"),
    key = paste(donor, time_interval, id))
                     
maternal_hits_key <- filter(maternal_hits, maternal == "maternal")$key

hits_by_maternal <- Mat_end_hits %>% 
  mutate(key = paste(donor, time_interval, id),
         maternal = ifelse(key %in% maternal_hits_key, "maternal", "endogenous"))
```
# Figure 6: MMR vaccination
## Prepare data

```{r}
# Identify vaccinated donors and prep data for virus
temp_donors <- Vaccination_Data %>%
  filter(species %in% c("Measles morbillivirus","Rubivirus rubellae")) %>%
  select(donor, cohort, Sex, Age_Years, Age_Months, Pathogen) %>%
  distinct() %>%
  group_by(donor) %>%
  summarise(n_vaccination = n(), .groups="drop") %>%
  select(donor) %>%
  distinct() %>%
  pull(donor) %>%
  unique()

temp_samples <- combined_metadata_dist %>% filter(donor %in% temp_donors)

# keep only donors with at least 3 sampled timepoints (true sample count)
temp_donors_initial <- temp_samples %>%
  distinct(donor, identifier) %>%
  count(donor, name = "n_samples") %>%
  filter(n_samples >= 3) %>%
  pull(donor) %>%
  unique()

# re-subset samples after filtering donors
temp_samples <- combined_metadata_dist %>% filter(donor %in% temp_donors_initial)

temp_peptides <- Viral_Annotations %>%
  filter(species %in% c("Measles morbillivirus","Rubivirus rubellae"))

# Define vax events (using measles as proxy for MMR)
temp_vaccination <- Vaccination_Data %>%
  filter(donor %in% temp_donors_initial, species %in% c("Measles morbillivirus")) %>%
  select(-c(species, Vaccine_Name_English)) %>%
  distinct() %>%
  group_by(donor) %>%
  arrange(Age_Years) %>%
  mutate(vaccine_dose = as.factor(row_number())) %>%
  ungroup()

# Process Z-scores and hits
temp_zs <- all_zs_combined %>%
  select(c(id, all_of(temp_samples$identifier))) %>%
  filter(id %in% temp_peptides$id) %>%
  bound_Z(upper = 15) %>%
  left_join(Viral_Annotations %>% select(c(id, Entry, `Entry name`, species, `Protein names`, Organism))) %>%
  pivot_longer(-c(id, Entry, `Entry name`, species, `Protein names`, Organism),
               names_to = "identifier", values_to = "Z_Bound") %>%
  mutate(hit = ifelse(Z_Bound >= 3.5, 1, 0)) %>%
  left_join(temp_samples %>% select(c(identifier, Age_Years, donor)))

#  Adaptive Filtering 

# Get Dose 1 age for windowing
dose1_age <- temp_vaccination %>%
  group_by(donor) %>%
  slice_min(order_by = as.numeric(as.character(vaccine_dose)), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(donor, dose1_age = Age_Years)

# Assess windows per donor
prepost_df <- temp_samples %>%
  left_join(dose1_age, by = "donor") %>%
  filter(!is.na(dose1_age)) %>%
  group_by(donor) %>%
  summarise(
    has_pre6  = any(Age_Years < dose1_age & Age_Years >= (dose1_age - 0.5), na.rm = TRUE),
    has_post6 = any(Age_Years > dose1_age & Age_Years <= (dose1_age + 0.5), na.rm = TRUE),
    n_samples = n_distinct(identifier),
    .groups = "drop"
  )

# Pick donor set based on data density
donors_pre_and_post <- prepost_df %>% filter(has_pre6 & has_post6) %>% pull(donor)
donors_pre_or_post  <- prepost_df %>% filter(has_pre6 | has_post6) %>% pull(donor)
donors_at_least3    <- prepost_df %>% filter(n_samples >= 3) %>% pull(donor)

if (length(donors_pre_and_post) > 0) {
  selected_donors <- donors_pre_and_post
} else if (length(donors_pre_or_post) > 0) {
  selected_donors <- donors_pre_or_post
} else {
  selected_donors <- donors_at_least3
}

# apply final filters to working sets
temp_donors      <- intersect(temp_donors_initial, selected_donors)
temp_samples     <- combined_metadata_dist %>% filter(donor %in% temp_donors)
temp_vaccination <- temp_vaccination %>% filter(donor %in% temp_donors)
temp_zs          <- temp_zs %>% filter(donor %in% temp_donors)

# Select Key Samples

# Birth sample
birth <- temp_samples %>%
  filter(Age_Years == 0) %>%
  group_by(donor) %>%
  slice_min(Age_Years, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(dose = "Birth") %>%
  select(donor, identifier, Age_Years, dose)

# Pre-vax sample
pre <- temp_samples %>%
  left_join(dose1_age, by = "donor") %>%
  filter(Age_Years < dose1_age) %>%
  mutate(dist_from_dose1 = abs(Age_Years - dose1_age)) %>%
  group_by(donor) %>%
  slice_min(dist_from_dose1, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  filter(Age_Years >= (6/12)) %>%
  mutate(dose = "Pre") %>%
  select(donor, identifier, Age_Years, dose)

# Post-dose samples
dose_ages <- temp_vaccination %>%
  mutate(dose_num = as.integer(as.character(vaccine_dose))) %>%
  select(donor, dose_num, dose_age = Age_Years)

post_candidates <- temp_samples %>%
  select(donor, identifier, Age_Years) %>%
  inner_join(dose_ages, by = "donor") %>%
  filter(Age_Years > dose_age) %>%
  mutate(dist_after = Age_Years - dose_age) %>%
  group_by(donor, dose_num) %>%
  slice_min(dist_after, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(dose = paste0("Dose ", dose_num)) %>%
  select(donor, identifier, Age_Years, dose)

# Combine and uniqueify
key_metadata_raw <- bind_rows(birth, pre, post_candidates) %>%
  mutate(dose_num = case_when(
    dose == "Birth" ~ 0L,
    dose == "Pre"   ~ 0L,
    str_detect(dose, "^Dose ") ~ as.integer(str_extract(dose, "\\d+")),
    TRUE ~ NA_integer_
  ))

max_dose <- key_metadata_raw %>% summarise(m = max(dose_num, na.rm = TRUE)) %>% pull(m)
dose_levels <- c("Birth", "Pre", paste0("Dose ", seq_len(max_dose)))

key_metadata <- key_metadata_raw %>%
  mutate(dose = factor(dose, levels = dose_levels)) %>%
  group_by(donor, identifier) %>%
  slice_max(order_by = as.integer(dose), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(donor, identifier, Age_Years, dose)

# final donor check for key samples count
temp_donors_final <- key_metadata %>%
  distinct(donor, identifier) %>%
  count(donor, name = "n_key_samples") %>%
  filter(n_key_samples >= 3) %>%
  pull(donor) %>%
  unique()

temp_donors_final <- intersect(temp_donors, temp_donors_final)
temp_samples      <- temp_samples %>% filter(donor %in% temp_donors_final)
key_metadata      <- key_metadata %>% filter(donor %in% temp_donors_final)
temp_zs           <- temp_zs %>% filter(donor %in% temp_donors_final)
```

# Additional processing for plot
```{r}
# Prep Plotting Objects 

key_zs <- temp_zs %>%
  filter(identifier %in% key_metadata$identifier) %>%
  left_join(key_metadata %>% select(c(identifier, dose)), by = "identifier") %>%
  mutate(dose = factor(dose, levels = dose_levels))

dose_labels_mmr <- key_metadata %>%
  group_by(donor, dose) %>%
  summarise(age = round(mean(Age_Years), 1), .groups = "drop") %>%
  mutate(
    dose = factor(dose, levels = dose_levels),
    dose_lab = paste0(as.character(dose), " (", age, "y)"),
    x_key = paste(donor, dose, sep = "___")
  ) %>%
  arrange(donor, dose)

key_zs2 <- key_zs %>%
  left_join(dose_labels_mmr %>% select(donor, dose, dose_lab, x_key),
            by = c("donor", "dose")) %>%
  mutate(x_key = factor(x_key, levels = unique(dose_labels_mmr$x_key)))

temp_peps <- key_zs2 %>%
  group_by(id) %>%
  summarise(n_hits = sum(hit), .groups = "drop") %>%
  filter(n_hits > 0)

# Define palettes
measles_pal <- colorRampPalette(c("gray98", "#F87E2E"))(101)
rubella_pal  <- colorRampPalette(c("gray98", "#FF36B0"))(101)

map_to_hex <- function(z, pal, maxZ = 15) {
  idx <- round((pmin(pmax(z, 0), maxZ) / maxZ) * 100) + 1
  pal[idx]
}

key_zs2 <- key_zs2 %>%
  mutate(
    MMRType = case_when(
      species == "Measles morbillivirus" ~ "Measles",
      species == "Rubivirus rubellae"   ~ "Rubella",
      TRUE ~ as.character(species)
    ),
    fill_hex = case_when(
      MMRType == "Measles" ~ map_to_hex(Z_Bound, measles_pal, maxZ = 15),
      MMRType == "Rubella" ~ map_to_hex(Z_Bound, rubella_pal, maxZ = 15),
      TRUE ~ map_to_hex(Z_Bound, measles_pal, maxZ = 15)
    )
  )
```



# Plot
```{r, fig.width=7,fig.height=6}
# Panel A: Heatmap
mmr_heat <- ggplot(key_zs2 %>% filter(id %in% temp_peps$id), aes(x = x_key, y = as.factor(id), fill = fill_hex)) +
  geom_tile() +
  facet_grid(MMRType ~ donor, scales = "free", space = "free") +
  scale_x_discrete(labels = setNames(dose_labels_mmr$dose_lab, dose_labels_mmr$x_key)) +
  scale_fill_identity(guide = "none") +
  ylab("VirScan Peptide") +
  theme_minimal() +
  theme(
    text = element_text(family = "Helvetica", color = "black", size = 7),
    axis.text.x = element_text(angle = 90, size = 7, vjust = 0.5, hjust = 1, color = "black", margin = margin(t = 0, b = 0)),
    axis.text.y = element_blank(),
    axis.title = element_text(size = 7),
    strip.text = element_text(size = 7),
    axis.ticks.y = element_blank(),
    legend.position = "none",
    panel.grid = element_blank(),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    strip.text.y.left = element_text(angle = 0),
    plot.margin = margin(t = 0, r = 0, b = 0, l = 0) # RESTORED
  ) +
  xlab("")

#7 Legends row 7
legend_df <- tibble(z = seq(0, 15, length.out = 101), meas_hex = measles_pal, rub_hex = rubella_pal, y = 1)
legend_breaks <- c(0, 5, 10, 15); legend_labels <- c("0", "5", "10", "15")

legend_measles <- ggplot(legend_df, aes(x = z, y = y, fill = meas_hex)) +
  geom_tile(height = 0.9) +
  annotate("text", x = 7.5, y = 1, label = "Measles", family = "Helvetica", size = 2.5,  color = "black") +
  scale_fill_identity() + scale_x_continuous(breaks = legend_breaks, labels = legend_labels, expand = c(0, 0)) +
  theme_minimal() + theme(text = element_text(family = "Helvetica", color = "black"), axis.title.x = element_blank(), axis.text.x = element_text(size = 7, color = "black"), axis.ticks.x = element_line(linewidth = 0.3, color = "black"), axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank(), plot.margin = margin(t = 0, r = 4, b = 0, l = 4))

legend_rubella <- ggplot(legend_df, aes(x = z, y = y, fill = rub_hex)) +
  geom_tile(height = 0.9) +
  annotate("text", x = 7.5, y = 1, label = "Rubella", family = "Helvetica", size = 2.5,  color = "black") +
  scale_fill_identity() + scale_x_continuous(breaks = legend_breaks, labels = legend_labels, expand = c(0, 0)) +
  theme_minimal() + theme(text = element_text(family = "Helvetica", color = "black"), axis.title.x = element_blank(), axis.text.x = element_text(size = 7, color = "black"), axis.ticks.x = element_line(linewidth = 0.3, color = "black"), axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank(), plot.margin = margin(t = 0, r = 4, b = 0, l = 4))

legend_title <- ggplot() + annotate("text", x = 0.5, y = 0.5, label = "Bounded Mean EBS", family = "Helvetica", size = 3, fontface = "plain", color = "black") + theme_void() + theme(plot.margin = margin(t = 0, r = 0, b = 0, l = 0))
top_legends <- cowplot::plot_grid(legend_measles + theme(plot.margin = margin(0, 5, 0, 0)), legend_rubella + theme(plot.margin = margin(0, 0, 0, 5)), ncol = 2, rel_widths = c(1, 1))
panel_top <- cowplot::plot_grid(legend_title, top_legends, ncol = 1, rel_heights = c(0.35, 1))
top_legends

# Plot Panel B 
temp_df_grouped <- temp_zs %>% group_by(donor, species, Age_Years) %>% summarise(n_hits_donor = sum(hit, na.rm = TRUE), .groups = "drop") %>% filter(donor %in% temp_donors_final)
vax_lines <- temp_vaccination %>% group_by(donor) %>% summarise(vax_age = first(Age_Years), .groups = "drop") %>% filter(donor %in% temp_donors_final)

p_mmr_hits_time <- ggplot(temp_df_grouped %>% mutate(species_label = case_when(species == "Measles morbillivirus" ~ "Measles", species == "Rubivirus rubellae" ~ "Rubella", TRUE ~ as.character(species))), aes(x = Age_Years, y = n_hits_donor, color = species_label)) +
  geom_line(linewidth = 0.8) + geom_point(size = 1.6) +
  geom_vline(data = vax_lines, aes(xintercept = vax_age), linetype = "dashed", linewidth = 0.6, color = "black", alpha = 0.6) +
  facet_wrap(~ donor, nrow = 1) +
  xlab("Age (years)") + ylab("# peptides targeted") +
  theme_minimal() +
  theme(text = element_text(family = "Helvetica"), axis.title = element_text(size = 7), strip.text = element_text(size = 7), axis.text = element_text(size = 7), legend.title = element_text(size = 7), legend.text = element_text(size = 7), legend.position = "top", legend.margin = margin(t = 0, r = 0, b = 0, l = 0), legend.box.margin = margin(t = -2, r = 0, b = -2, l = 0), panel.grid = element_blank(), panel.border = element_rect(color = "black", fill = NA), plot.margin = margin(t = 0, r = 0, b = 0, l = 5)) +
  scale_color_manual(name = "Virus", values = c("Measles" = pal1[1], "Rubella" = pal1[8]))

# Final Combine 
panel_top_conv <- ggplotify::as.ggplot(panel_top) + theme(plot.margin = margin(t = 0, r = 0, b = 0, l = 0))

Fig_6 <- ggarrange(
  ggarrange(ggplot(NULL) + theme_void(), panel_top_conv, ggplot(NULL) + theme_void(), widths = c(0.75, 18, 0.75), nrow = 1),
  ggarrange(ggplot(NULL) + theme_void(), mmr_heat, widths = c(0.6, 18), nrow = 1),
  ggarrange(p_mmr_hits_time, ggplot(NULL) + theme_void(), widths = c(18, 0.65), nrow = 1),
  labels = c("a", "", "b"), ncol = 1, heights = c(0.18, 1.25, 0.8), font.label = list(family = "Helvetica", size = 8)
)

Fig_6
ggsave(file=paste0(folder_path_save, "/AVARS_Figure_6_Final.pdf"))
dev.off()
```

# Figure S11: Influenza
# Influenza analysis
# Process data for downstream analysis
```{r, fig.height=6}
# Identify vaccinated donors and prep data for virus
temp_donors <- Vaccination_Data %>% filter(species %in% c("Influenza B virus","Influenza A virus")) %>% select(donor, cohort, Sex, Age_Years, Age_Months, Pathogen) %>% distinct()%>% group_by(donor)%>% select(donor) %>% distinct() %>% unlist() %>% unname()
temp_samples <- combined_metadata_dist %>% filter(donor %in% temp_donors)
temp_peptides <- Viral_Annotations %>% filter(species %in% c("Influenza A virus","Influenza B virus"))
temp_vaccination <- Vaccination_Data %>% filter(donor %in% temp_donors, species %in% c("Influenza A virus","Influenza B virus")) %>% select(-c(species,Vaccine_Name_English)) %>% distinct() %>% group_by(donor) %>% arrange(Age_Years)%>% mutate(vaccine_dose=as.factor(row_number())) 
temp_consec_zs <- Birth_Consec_Zs %>% filter(donor %in% temp_donors, id %in% temp_peptides$id)
temp_zs <- all_zs_combined %>% select(c(id,all_of(temp_samples$identifier))) %>% filter(id %in% temp_peptides$id) %>% bound_Z(upper=25) %>% left_join(Viral_Annotations %>% select(c(id,Entry,`Entry name`,species,`Protein names`,Organism))) %>% pivot_longer(-c(id,Entry,`Entry name`,species,`Protein names`,Organism),names_to = "identifier",values_to = "Z_Bound") %>% mutate(hit=ifelse(Z_Bound>=3.5,1,0)) %>% left_join(temp_samples %>% select(c(identifier,Age_Years,donor)))
```

# Prepare data and identify which flu strains to include in plots
Strains vaccinated against in window "Influenza A virus (A/Brisbane/59/2007(H1N1))", "Influenza A virus (A/Brisbane/10/2007(H3N2))","Influenza A virus (A/California/07/2009(H1N1))", "Influenza B virus (B/Brisbane/60/2008)"

```{r, fig.width=10}
# Limit to human influenza strains
influenza_peptides <- Viral_Annotations %>% 
  filter(species %in% c("Influenza A virus","Influenza B virus")) %>%
  filter(!grepl("swine",Organism,ignore.case=TRUE),
         !grepl("chicken",Organism,ignore.case=TRUE),
         !grepl("mallard",Organism,ignore.case=TRUE),
         !grepl("ruddy turnstone",Organism,ignore.case=TRUE),
         !grepl("duck",Organism,ignore.case=TRUE),
         !grepl("Turkey",Organism,ignore.case=TRUE),
         !grepl("Equine",Organism,ignore.case=TRUE),
         !grepl("Seal",Organism,ignore.case=TRUE),
         !grepl("camel",Organism,ignore.case=TRUE),
         !grepl("teal",Organism,ignore.case=TRUE),
         !grepl("goose",Organism,ignore.case=TRUE),
         !grepl("fowl",Organism,ignore.case=TRUE))

# Subset maternal/endogenous labeled df to these peptide responses only
temp_hits_by_maternal <- hits_by_maternal %>% filter(id %in% influenza_peptides$id, identifier %in% temp_samples$identifier) %>% select(c(id,species,Organism,identifier,hit,donor,Age_Years,zscore,trunc_zscore,maternal)) %>% left_join(IgG_Data %>% select(c(identifier,collect_date))) %>% mutate(donor_peptide_key=paste0(donor,"_",id))
temp_ever_hit <- temp_hits_by_maternal %>% group_by(donor_peptide_key,id,species,Organism) %>% summarise(n_hits=sum(hit)) %>% filter(n_hits>0)
temp_hits_by_maternal %>% select(id) %>% distinct() %>% nrow() #3935 peptides total
temp_ever_hit %>% ungroup()%>% select(id) %>% distinct() %>% nrow() #1766 flu peptides hit

#count how many peptides per entry are in the library, how many birth samples target the entry, and how many samples closest to 2 years old target the sample
flu_entries <- Viral_Annotations %>% filter(id %in% influenza_peptides$id) %>% 
  group_by(Entry,`Entry name`,NCBI_txid,genus,species,`Protein names`,`Gene names`,`Taxonomic lineage (ALL)`,Organism) %>% 
  summarise(n_peptides_in_lib=n()) %>% # count how many peptides per entry are in the library
  left_join(temp_hits_by_maternal %>% 
              left_join(Viral_Annotations %>% select(c(id,Entry))) %>% 
              filter(Age_Years==0,hit==1) %>% group_by(donor,Organism,Entry) %>% 
              summarise(n_pep_targ=sum(hit)) %>% ungroup %>% group_by(Organism,Entry) %>% 
              summarise(n_rec_entry_maternal = n(), avg_pep_rec_maternal=mean(n_pep_targ))
            ) %>% # how many birth samples target the entry (at least 1 peptide)
  left_join(temp_hits_by_maternal %>% group_by(id,donor) %>% 
              mutate(dist_from_2=abs(Age_Years-2)) %>% 
              slice_min(order_by = dist_from_2,n=1) %>% 
              ungroup() %>% left_join(Viral_Annotations %>% select(c(id,Entry))) %>%
              filter(hit==1)  %>% group_by(donor,Organism,Entry) %>% 
              summarise(n_pep_targ=sum(hit)) %>% group_by(Organism,Entry) %>% 
              summarise(n_rec_entry_endog = n(), avg_pep_rec_endog=mean(n_pep_targ))
            ) # how many samples closest to 2 years old target the entry (at least 1 peptide)

# Limit to entries recognized and label HA and N
temp_ever_hit_entries <- temp_ever_hit %>% ungroup() %>% select(id) %>% distinct() %>% left_join(influenza_peptides %>% select(c(id,Entry)))
```
# Identify and label peptides and proteins for strains of interest
```{r}
# Focus on Influenza A
flu_a_entries <- flu_entries %>% 
  filter(species=="Influenza A virus") %>% 
  mutate(n_rec_entry_maternal=ifelse(is.na(n_rec_entry_maternal),0,n_rec_entry_maternal)) %>% 
  mutate(n_rec_entry_endog=ifelse(is.na(n_rec_entry_endog),0,n_rec_entry_endog)) %>% 
  filter(Entry %in% temp_ever_hit_entries$Entry) %>% 
  mutate(`Gene names`=ifelse(grepl("Neuraminidase",`Protein names`,ignore.case=TRUE),"NA",`Gene names`)) %>% 
  mutate(`Gene names`=ifelse(grepl("Hemagglutinin",`Protein names`,ignore.case=TRUE),"HA",`Gene names`))

flu_a_entries

# Identify strains with NA and HA sequences
HA_NA_entries <- flu_a_entries %>% 
  filter(`Gene names` %in% c("HA","NA"), n_peptides_in_lib > 4) %>% 
  mutate(
    is_pdm09 = grepl("2009", Organism),
    plot_strain_label = ifelse(is_pdm09,
                              "Influenza A virus (A(H1N1)pdm09 lineage)",
                              Organism),
    score = pmax(avg_pep_rec_maternal, avg_pep_rec_endog, na.rm = TRUE)
  ) %>%
  group_by(plot_strain_label, `Gene names`) %>%
  filter(!any(is_pdm09) | row_number(desc(score)) == 1) %>%  # only collapse pdm09 groups
  ungroup() %>%
  select(-is_pdm09, -score)

HA_NA_entries %>% group_by(plot_strain_label) %>% summarise(n_HA_ent=sum(`Gene names`=="HA"),n_NA_ent=sum(`Gene names`=="NA")) %>% filter(n_HA_ent>0,n_NA_ent>0) %>% select(plot_strain_label)  # of this list of strains with NA and HA sequences in the library, "Influenza A virus (strain A/Brevig Mission/1/1918 H1N1) (Influenza A virus (strain A/South Carolina/1/1918 H1N1))","Influenza A virus (strain A/Korea/426/1968 H2N2)","Influenza A virus (strain A/Hong Kong/1/1968 H3N2)","Influenza A virus (strain A/Victoria/3/1975 H3N2)","Influenza A virus (strain A/Beijing/353/1989 H3N2)","Influenza A virus (A/Brisbane/10/2007(H3N2))","Influenza A virus (strain A/Kiev/59/1979 H1N1)","Influenza A virus (strain A/China:Nanchang/11/1996 H1N1)","Influenza A virus (strain A/USA:Texas/UR06-0195/2007 H1N1)","Influenza A virus (A(H1N1)pdm09 lineage)","Influenza A virus (strain A/Hong Kong/212/2003 H5N1 genotype Z+)" are from distinct antigenic clusters

HA_NA_entries <- HA_NA_entries %>% filter(plot_strain_label %in% c("Influenza A virus (strain A/Brevig Mission/1/1918 H1N1) (Influenza A virus (strain A/South Carolina/1/1918 H1N1))","Influenza A virus (strain A/Korea/426/1968 H2N2)","Influenza A virus (strain A/Hong Kong/1/1968 H3N2)","Influenza A virus (strain A/Victoria/3/1975 H3N2)","Influenza A virus (strain A/Beijing/353/1989 H3N2)","Influenza A virus (A/Brisbane/10/2007(H3N2))","Influenza A virus (strain A/Kiev/59/1979 H1N1)","Influenza A virus (strain A/China:Nanchang/11/1996 H1N1)","Influenza A virus (strain A/USA:Texas/UR06-0195/2007 H1N1)","Influenza A virus (A(H1N1)pdm09 lineage)","Influenza A virus (strain A/Hong Kong/212/2003 H5N1 genotype Z+)"))

# Determine who is flu exposed.
temp_donors <- Vaccination_Data %>% 
  filter(species %in% c("Influenza B virus","Influenza A virus")) %>% 
  select(donor, cohort, Sex, Age_Years, Age_Months, Pathogen) %>% 
  distinct() %>% 
  group_by(donor) %>% 
  summarise(n_vax=n()) %>%
  filter(n_vax>1) %>% 
  select(donor) %>% 
  distinct() %>% 
  unlist() %>% 
  unname()

hm_hits <- Mat_end_hits %>% 
  filter(donor %in% temp_donors) %>% 
  left_join(Viral_Annotations %>% select(c(Entry, id))) %>% 
  filter(Entry %in% HA_NA_entries$Entry) %>% 
  left_join(HA_NA_entries %>% ungroup() %>% select(c(Entry,plot_strain_label,`Gene names`))) %>% 
  mutate(HA_number = str_extract(plot_strain_label, "(?<=H)\\d+"), NA_number = str_extract(plot_strain_label, "(?<=N)\\d+"), HA_NA_subtype = str_extract(plot_strain_label, "H\\d+N\\d+")) # adds substype information

# add shortened lables for plot
temp<-data.frame("plot_strain_label"=c(unique(hm_hits$plot_strain_label)),"short_strain"=c("A/H1N1pdm09/2009","A/Brevig Mission/1/1918(H1N1)","A/Hong Kong/212/2003(H5N1)","A/Texas/UR06-0195/2007(H1N1)","A/Kiev/59/1979(H1N1)","A/Korea/426/1968(H2N2)","A/Victoria/3/1975(H3N2)","A/China:Nanchang/11/1996(H1N1)","A/Beijing/353/1989(H3N2)","A/Hong Kong/1/1968(H3N2)","A/Brisbane/10/2007(H3N2)"))%>%
  mutate(
    short_strain_pretty2 = case_when(
      
      # pandemic strain
      str_detect(short_strain, "pdm09") ~ "A/pdm09/2009 (H1N1)",
      
      # general case: A/Location/Number/Year(HxNy)
      TRUE ~ str_replace(
        short_strain,
        "A/([^/]+)/[^/]+/([0-9]{4}).*\\((H[0-9]N[0-9]).*",
        "A/\\1/\\2 (\\3)"
      )
    ),
    
    # remove country prefix before colon (e.g., China:Nanchang → Nanchang)
    short_strain_pretty2 = str_replace(short_strain_pretty2,
                                       "A/[^:]+:([^/]+)/",
                                       "A/\\1/"),
    
    # allow wrapping at slash and before subtype
    short_strain_wrap = short_strain_pretty2 %>%
      str_replace_all("/", "/ ") %>%
      str_replace_all("\\(", " (")
  )

hm_hits <-hm_hits %>% left_join(temp)

rm(temp)

# Define key samples (birth and immediate after last vaccine dose)
temp_samples <- combined_metadata_dist %>% 
  filter(donor %in% temp_donors) 

temp_peptides <-unique(hm_hits$id) 

temp_consec_zs <- Birth_Consec_Zs %>% filter(donor %in% temp_donors,id %in% temp_peptides) 

key_metadata <- temp_samples %>% filter(Age_Years==0) %>% mutate(dose="Birth") %>% bind_rows(temp_samples %>% left_join(temp_vaccination %>% group_by(donor) %>% slice_max(order_by = vaccine_dose,n=1) %>% mutate(dose_max_age=Age_Years) %>% select(c(donor,dose_max_age,vaccine_dose))) %>% filter(Age_Years>dose_max_age) %>% mutate(dist_from_dose_4=Age_Years-dose_max_age) %>% group_by(donor) %>% arrange(dist_from_dose_4) %>% slice_min(dist_from_dose_4,n=1) %>% select(-c(dose_max_age,dist_from_dose_4)) %>% mutate(dose="Last Dose")) 

```

# Label donor data and define donor subsets
```{r}
dose1_age <- temp_vaccination %>%
  group_by(donor) %>%
  slice_min(order_by = as.numeric(as.character(vaccine_dose)), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(donor, dose1_age = Age_Years)


last_dose_age <- temp_vaccination %>%
  group_by(donor) %>%
  slice_max(order_by = as.numeric(as.character(vaccine_dose)), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(donor, last_dose_age = Age_Years)

birth <- temp_samples %>%
  filter(Age_Years == 0) %>%
  group_by(donor) %>%
  slice_min(Age_Years, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(dose = "Birth") %>%
  select(donor, identifier, Age_Years, dose)

# Sample before vaccination but after at least 6 months of age
pre <- temp_samples %>%
  left_join(dose1_age, by = "donor") %>%
  filter(Age_Years < dose1_age) %>%
  mutate(dist_from_dose1 = abs(Age_Years - dose1_age)) %>%
  group_by(donor) %>%
  slice_min(dist_from_dose1, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  filter(Age_Years >= (6/12)) %>%  # >= 6 months
  mutate(dose = "Pre") %>%
  select(donor, identifier, Age_Years, dose)

# Sample after but closest to last vaccine dose
last <- temp_samples %>%
  left_join(last_dose_age, by = "donor") %>%
  filter(Age_Years > last_dose_age) %>%
  mutate(dist_from_last = Age_Years - last_dose_age) %>%
  group_by(donor) %>%
  slice_min(dist_from_last, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(dose = "Last") %>%
  select(donor, identifier, Age_Years, dose)

key_metadata2 <- bind_rows(birth, pre, last) %>%
  mutate(dose = factor(dose, levels = c("Birth","Pre","Last")))

# keep donors with all 3 timepoints
keep_donors <- key_metadata2 %>%
  count(donor, dose) %>%
  count(donor) %>%
  filter(n == 3) %>%
  pull(donor)

key_metadata2 <- key_metadata2 %>% filter(donor %in% keep_donors)
```


# Figure S11
```{r, fig.width=7, fig.height=3.5}
# Set Parameters
flu_seasons <- tibble(xmin = as.Date(c("2008-11-01","2009-11-01","2010-11-01","2011-11-01","2012-11-01")), xmax = as.Date(c("2009-04-30","2010-04-30","2011-04-30","2012-04-30","2013-04-30")), fill_lab = "Flu season") # Define flu season to November 1 to April 15 of each year
# Events dataframe (blood draws + vaccinations) 
blood_draws <- IgG_Data %>% filter(identifier %in% temp_samples$identifier) %>% transmute( donor, date = as.Date(collect_date, format = "%m/%d/%y"), event_row = "Blood draw", event_color = "Blood draw" ) %>% distinct() 
vax_events <- temp_vaccination %>% transmute(donor, date = do_vax, event_row = ifelse(grepl("H1N1", Vaccine_Targets, ignore.case = TRUE), "Pandemic Vaccine", "Seasonal Vaccine"), event_color = event_row ) %>% distinct() 
event_df <- bind_rows(blood_draws, vax_events) %>% mutate(event_row = factor(event_row, levels = c("Blood draw","Seasonal Vaccine","Pandemic Vaccine"))) 
# Filter hits to those peptides ever hit by a given donor and flu strains of interest then group by strain and sum 
hits_df <- temp_hits_by_maternal %>% 
  filter(donor_peptide_key %in% temp_ever_hit$donor_peptide_key) %>% 
  mutate(plot_strain_label=ifelse(Organism %in% c("Influenza A virus (A/New York/19/2009(H1N1))", "Influenza A virus (A/Ohio/07/2009(H1N1))"), "Influenza A virus (A(H1N1)pdm09 lineage)", Organism)) %>% # All 2009 strains represent the pandemic strain. These two entries are HA and NA entries with good coverage
  filter (plot_strain_label %in% c( "Influenza A virus (A/Brisbane/59/2007(H1N1))", "Influenza A virus (A/Brisbane/10/2007(H3N2))", "Influenza A virus (A(H1N1)pdm09 lineage)", "Influenza B virus (B/Brisbane/60/2008)","Influenza A virus (strain A/Bangkok/1/1979 H3N2)","Influenza A virus (strain A/Beijing/352/1989 H3N2)")) %>% # These are vaccine strains administered during sampling and pre-birth strains from different antigenic clusters
  group_by(identifier, donor, Age_Years, maternal, collect_date, species, plot_strain_label) %>% 
  summarise(n_hits = sum(hit), .groups = "drop") %>% 
  mutate(date = as.Date(collect_date, format = "%m/%d/%y")) %>% mutate(
  plot_strain_label_pretty2 = case_when(
    # pdm09 collapse
    plot_strain_label == "Influenza A virus (A(H1N1)pdm09 lineage)" ~ "A/pdm09/2009 (H1N1)",

    # Influenza A "A/Location/Number/Year(Subtype)" -> "A/Location/Year (Subtype)"
    str_detect(plot_strain_label, "^Influenza A virus \\(A/") ~
      plot_strain_label %>%
        str_replace("^Influenza A virus \\(", "") %>%
        str_replace("\\)$", "") %>%
        str_replace("A/([^/]+)/[^/]+/([0-9]{4})\\((H[0-9]N[0-9])\\)", "A/\\1/\\2 (\\3)"),

    # Influenza A "strain A/Location/Number/Year HxNy" -> "A/Location/Year (HxNy)"
    str_detect(plot_strain_label, "^Influenza A virus \\(strain A/") ~
      plot_strain_label %>%
        str_replace("^Influenza A virus \\(strain ", "") %>%
        str_replace("\\)$", "") %>%
        str_replace("A/([^/]+)/[^/]+/([0-9]{4}) (H[0-9]N[0-9])", "A/\\1/\\2 (\\3)"),

    # Influenza B: keep tidy
    plot_strain_label == "Influenza B virus (B/Brisbane/60/2008)" ~ "B/Brisbane/2008",

    TRUE ~ plot_strain_label
  )
)

# Split donors into two groups for plotting 
donor_levels <- sort(intersect(unique(event_df$donor), unique(hits_df$donor))) 
half <- ceiling(length(donor_levels) / 2) 
donors_1 <- donor_levels[1:half] 
donors_2 <- donor_levels[(half + 1):length(donor_levels)]

  x_min <- min(c(hits_df$date, event_df$date, flu_seasons$xmin), na.rm = TRUE)
  x_max <- max(c(hits_df$date, event_df$date, flu_seasons$xmax), na.rm = TRUE)

make_combined_plot <- function(donor_subset) {
  event_sub <- event_df %>%
    filter(donor %in% donor_subset) %>%
    mutate(donor = factor(donor, levels = donor_subset))
  hits_sub <- hits_df %>%
    filter(donor %in% donor_subset) %>%
    mutate(donor = factor(donor, levels = donor_subset))
  
  event_plot <- ggplot(event_sub, aes(x = date, y = event_row)) +
    geom_rect(data = flu_seasons, inherit.aes = FALSE,
              aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = fill_lab),
              alpha=0.15) +
    geom_point(aes(colour = event_color), pch=20, size=1) +
    facet_wrap(~donor, nrow=1, scales = "free_x") +
    scale_fill_manual(values = c("Flu season" = pal1[3]), name = "") +
    scale_color_manual(values = c("Blood draw" = "black",
                                  "Seasonal Vaccine" = pal1[4],
                                  "Pandemic Vaccine" = pal1[7]),
                       name = "Event") +
    coord_cartesian(xlim = c(x_min, x_max)) +
    theme_minimal() +
    theme(text             = element_text(family = "Helvetica", color = "black", size = 7),
          axis.title.x     = element_text(size=7),
          axis.title.y     = element_blank(),
          axis.text.y        = element_text(size = 7, color = "black"),
          axis.text.x        = element_text(size = 7, angle=90, color = "black", hjust = 0, vjust = 0.5),
          axis.ticks.x = element_line(),
          legend.position  = "none",    
          strip.background = element_blank(),
          strip.text       = element_blank(),
          panel.spacing    = unit(6, "pt"),
          panel.grid       = element_blank(),
          panel.border     = element_rect(fill=NA,colour = "black")
          ) +
    guides(fill = "none", colour = "none") +
    xlab("Date") 
  
  
  hit_plot <- ggplot(hits_sub, aes(x = date, y = n_hits)) +
    geom_rect(data = flu_seasons, inherit.aes = FALSE,
              aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = fill_lab),
              alpha=0.15) +
    geom_line(aes(color = plot_strain_label_pretty2, linetype = maternal)) +
    facet_wrap(~ donor, nrow = 1, scales = "free_x") +
    scale_fill_manual(values = c("Flu season" = pal1[3]), name = "") +
    theme_minimal() +
    coord_cartesian(xlim = c(x_min, x_max)) +
    theme(text             = element_text(family = "Helvetica", color = "black", size = 7),
          axis.title.y     = element_text(size=7),
          axis.title.x     = element_blank(),
          axis.text.y      = element_text(size = 7, color = "black"),
          axis.text.x      = element_blank(),
          strip.text       = element_text(size = 7),
          panel.spacing    = unit(6, "pt"),
          legend.position  = "none",
          panel.grid       = element_blank(),
          panel.border     = element_rect(fill=NA,colour = "black"),
          legend.title     = element_text(size = 7),
          legend.text      = element_text(size = 7),
          legend.box       = "vertical"
          ) +
    ylab("Number of hits") +
    scale_color_manual(values = pal1[c(1,9,4,5,7,3)], name = "Strain") +
    scale_linetype_discrete(name = "Antibody source")   +
    ylim(c(0,15))
  
    ggpubr::ggarrange(
    hit_plot + theme(plot.margin = margin(0, 5.5, 0, 5.5)),
    event_plot + theme(plot.margin = margin(0, 5.5, 0, 5.5)),
    ncol = 1,
    heights = c(1.5, 1.25),
    common.legend = FALSE,
    legend = "none",
    align = "v"
  )
}


legend_grob <- ggpubr::get_legend(
  ggplot(
    hits_df %>%
      filter(donor %in% donors_1) %>%
      mutate(donor = factor(donor, levels = donors_1)),
    aes(x = date, y = n_hits)
  ) +
  geom_rect(data = data.frame(dummy = 1),inherit.aes = FALSE, aes(xmin=as.Date("2011-11-1"),xmax =as.Date("2012-4-15"),fill = "Flu season", 
            ymin=0, ymax = 17),
            alpha=0.15)  +
    # real legend mappings
    geom_line(aes(color = plot_strain_label_pretty2, linetype = maternal)) +
    scale_color_manual(values = pal1[c(1,9,4,5,7,3)], name = "Strain") +
    scale_linetype_discrete(name = "Antibody source") +
    scale_fill_manual(values = c("Flu season" = pal1[3]), name = "") +
    theme_minimal() +
    theme(
      legend.position = "top",
      legend.box = "horizontal",
      legend.direction = "vertical",
      legend.title = element_text(size = 7),
      legend.text  = element_text(size = 7),
      plot.margin = margin(0, 0, 0, 0)
    ) +
    guides(
      fill = guide_legend(order = 1, override.aes = list(alpha = 0.15, colour = NA)),
      linetype = guide_legend(order = 2),
      colour = guide_legend(order = 3, ncol=2)
    )
)


Fig_S11_A <- ggpubr::ggarrange(
  ggpubr::as_ggplot(legend_grob),
    make_combined_plot(keep_donors),
  nrow = 2,
  heights = c(0.4, 0.8)
)
Fig_S11_A
```
# Generate heatmap of number of peptides targeted for key strains
```{r, fig.width=7}
# Key timepoints for heatmap: Birth, Pre-vaccination (>=9 months), and first sample after last dose
# Summarize for each donor
heat_df <- hm_hits %>%
  filter(donor %in% keep_donors, identifier %in% key_metadata2$identifier) %>%
  left_join(key_metadata2 %>% select(identifier, dose), by = "identifier") %>%
  group_by(donor, short_strain, `Gene names`, dose) %>%
  summarise(
    mean_z = mean(zscore, na.rm = TRUE),
    n_hits = sum(hit == 1, na.rm = TRUE),
    .groups = "drop"
  )

strain_levels <- c(
  "A/Brevig Mission/1/1918(H1N1)",
  "A/Kiev/59/1979(H1N1)",
  "A/China:Nanchang/11/1996(H1N1)",
  "A/Texas/UR06-0195/2007(H1N1)",
  "A/H1N1pdm09/2009",
  "A/Korea/426/1968(H2N2)",
  "A/Hong Kong/1/1968(H3N2)",
  "A/Victoria/3/1975(H3N2)",
  "A/Beijing/353/1989(H3N2)",
  "A/Brisbane/10/2007(H3N2)",
  "A/Hong Kong/212/2003(H5N1)"
) # Order by subtype and year



# Make a donor-specific x key so ordering can differ by facet (even with free_x)
dose_labels <- key_metadata2 %>%
  group_by(donor, dose) %>%
  summarise(age = round(mean(Age_Years), 1), .groups = "drop") %>%
  mutate(
    dose = factor(dose, levels = c("Birth","Pre","Last")),
    dose_lab = paste0(as.character(dose), " (", age, "y)"),
    x_key = paste(donor, dose, sep = "___")   # donor-specific x
  ) %>%
  arrange(donor, dose)

heat_df2 <- heat_df %>%
  left_join(dose_labels %>% select(donor, dose, dose_lab, x_key),
            by = c("donor","dose")) %>%
  mutate(
    x_key = factor(x_key, levels = unique(dose_labels$x_key))  # global order = donor then Birth/Pre/Last
  )

Fig_S11_B <- ggplot(
  heat_df2,
  aes(
    x = x_key,
    y = factor(short_strain, levels = strain_levels),
    fill = as.factor(n_hits)
  )
) +
  geom_tile(color = "white", linewidth = 0.25) +
  facet_grid(`Gene names` ~ donor, scales = "free_x", space = "free_x") +
  scale_x_discrete(labels = setNames(dose_labels$dose_lab, dose_labels$x_key)) +
  scale_fill_manual(
    values = c("white", "#E6E0FB", "#CDC2F7", "#B3A3F2", "#9870F5", "#5B2CBF", "#2E1560"),
    name = "# peptide\nhits"
  ) +
  theme_minimal() +
  theme(
    text = element_text(family = "Helvetica", colour = "black", size = 7),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 7),
    axis.text.y = element_text(size = 7),
    strip.text.x = element_text(size = 7),
    strip.text.y = element_text(size = 7),
    legend.title = element_text(size = 7),
    legend.text  = element_text(size = 7),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "grey99")
  ) +
  xlab("") +
  ylab("Strain")

Fig_S11_B
```

```{r,fig.width=7,fig.height=6.5}
ggarrange(ggarrange(Fig_S11_A,ggplot(NULL)+theme_void(),ncol = 2,widths = c(15,1)),
          Fig_S11_B +
            theme(legend.background = element_rect(color = "black"),
                  legend.position = "right",
                  axis.text = element_text(color = "black")) +
            guides(fill = guide_legend(ncol=1)),
          labels=c("a","b"),
          ncol=1,
          font.label = list(family="Helvetica",size=8),
          heights = c(2.2,2.5)
          )
ggsave(file=paste0(folder_path_save, "/AVARS_Figure_S11_Final.pdf"))
dev.off()

```


# Figure S12: Polio and Rubella
```{r}
# Polio
# Identify vaccinated donors and prep data for virus 
temp_donors <- Vaccination_Data %>% 
  filter(species=="Enterovirus C",Age_Years<1.25) %>% 
  group_by(donor) %>% 
  summarise(n_vaccination = n()) %>% 
  filter(n_vaccination==3) %>% 
  select(donor) %>% 
  distinct() %>% 
  unlist() %>% 
  unname() 
temp_samples <- combined_metadata_dist %>% filter(donor %in% temp_donors) 

temp_peptides <- Viral_Annotations %>% 
  filter(species=="Enterovirus C") %>% 
  filter(grepl("Poliovirus",Organism)) 

temp_vaccination <- Vaccination_Data %>% 
  filter(donor %in% temp_donors, species=="Enterovirus C") %>% 
  group_by(donor) %>% 
  arrange(Age_Years) %>% 
  mutate(vaccine_dose=as.factor(row_number())) 

temp_consec_zs <- Birth_Consec_Zs %>% filter(donor %in% temp_donors,species=="Enterovirus C") 

temp_zs <- all_zs_combined %>% 
  select(c(id,all_of(temp_samples$identifier))) %>% 
  filter(id %in% temp_peptides$id) %>% 
  bound_Z(upper=15) %>% 
  left_join(Viral_Annotations %>% select(c(id,Entry,`Entry name`,species,`Protein names`,Organism))) %>% 
  pivot_longer(-c(id,Entry,`Entry name`,species,`Protein names`,Organism), names_to = "identifier",values_to = "Z_Bound") %>% 
  mutate(hit=ifelse(Z_Bound>=3.5,1,0)) %>% 
  left_join(temp_samples %>% select(c(identifier,Age_Years,donor))) 

temp_zs <- temp_zs %>% 
  mutate(Poliovirus=ifelse(grepl("type 1",Organism),"Type 1",NA)) %>% 
  mutate(Poliovirus=ifelse(grepl("type 2",Organism), "Type 2",Poliovirus)) %>% 
  mutate(Poliovirus=ifelse(grepl("type 3",Organism),"Type 3",Poliovirus)) # label by polio type

# Key timepoints: Birth, Pre (closest < Dose 1; >=6mo), and first sample after each dose 
# Key timepoints: Birth, Pre (closest < Dose 1; >=6mo), and first sample after each dose 

# dose ages
dose1_age <- temp_vaccination %>%
  group_by(donor) %>%
  slice_min(order_by = as.numeric(as.character(vaccine_dose)), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(donor, dose1_age = Age_Years)

dose2_age <- temp_vaccination %>%
  filter(vaccine_dose == 2) %>%
  transmute(donor, dose2_age = Age_Years)

dose3_age <- temp_vaccination %>%
  filter(vaccine_dose == 3) %>%
  transmute(donor, dose3_age = Age_Years)

# Birth
birth <- temp_samples %>%
  filter(Age_Years == 0) %>%
  group_by(donor) %>%
  slice_min(Age_Years, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(dose = "Birth") %>%
  select(donor, identifier, Age_Years, dose)

# Pre-vax: closest sample before Dose 1
pre <- temp_samples %>%
  left_join(dose1_age, by = "donor") %>%
  filter(Age_Years < dose1_age) %>%
  mutate(dist_from_dose1 = abs(Age_Years - dose1_age)) %>%
  group_by(donor) %>%
  slice_min(dist_from_dose1, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(dose = "Pre") %>%
  select(donor, identifier, Age_Years, dose)

# Post-dose 1
post_dose1 <- temp_samples %>%
  left_join(dose1_age, by = "donor") %>%
  filter(Age_Years > dose1_age) %>%
  mutate(dist_from_dose1 = Age_Years - dose1_age) %>%
  group_by(donor) %>%
  slice_min(dist_from_dose1, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(dose = "Dose 1") %>%
  select(donor, identifier, Age_Years, dose)

# Post-dose 2
post_dose2 <- temp_samples %>%
  left_join(dose2_age, by = "donor") %>%
  filter(Age_Years > dose2_age) %>%
  mutate(dist_from_dose2 = Age_Years - dose2_age) %>%
  group_by(donor) %>%
  slice_min(dist_from_dose2, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(dose = "Dose 2") %>%
  select(donor, identifier, Age_Years, dose)

# Post-dose 3
post_dose3 <- temp_samples %>%
  left_join(dose3_age, by = "donor") %>%
  filter(Age_Years > dose3_age) %>%
  mutate(dist_from_dose3 = Age_Years - dose3_age) %>%
  group_by(donor) %>%
  slice_min(dist_from_dose3, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(dose = "Dose 3") %>%
  select(donor, identifier, Age_Years, dose)

key_metadata <- bind_rows(birth, pre, post_dose1, post_dose2, post_dose3) %>%
  mutate(dose = factor(dose, levels = c("Birth","Pre","Dose 1","Dose 2","Dose 3")))

key_metadata <- key_metadata %>%
  group_by(donor, identifier) %>%
  # slice_max on integer value of dose 
  slice_max(order_by = as.integer(dose), n = 1, with_ties = FALSE) %>%
  ungroup()

# Filter
key_zs <- temp_zs %>%
  filter(identifier %in% key_metadata$identifier) %>%
  left_join(key_metadata %>% select(identifier, dose), by = "identifier")

# Add age labels
# Create donor-specific x labels: "Dose (age y)"
dose_labels_polio <- key_metadata %>%
  group_by(donor, dose) %>%
  summarise(age = round(mean(Age_Years), 1), .groups = "drop") %>%
  mutate(
    dose = factor(dose, levels = c("Birth","Pre","Dose 1","Dose 2","Dose 3")),
    dose_lab = paste0(as.character(dose), " (", age, "y)"),
    x_key = paste(donor, dose, sep = "___")
  ) %>%
  arrange(donor, dose)

# Attach to key_zs
key_zs <- key_zs %>%
  left_join(dose_labels_polio %>% select(donor, dose, dose_lab, x_key),
            by = c("donor","dose")) %>%
  mutate(
    x_key = factor(x_key, levels = unique(dose_labels_polio$x_key))
  )

# Plot 
temp_peps <- key_zs %>%
  group_by(id) %>%
  summarise(n_hits = sum(hit)) %>%
  filter(n_hits > 0)

Fig_S12_A <- 
  ggplot(
  key_zs %>% filter(id %in% temp_peps$id),
  aes(x = x_key, y = as.factor(id), fill = Z_Bound)
) +
  geom_tile() +
  facet_grid(Poliovirus ~ donor, scales = "free", space = "free") +
  scale_x_discrete(
    labels = setNames(dose_labels_polio$dose_lab, dose_labels_polio$x_key)
  ) +
  scale_fill_gradientn(
    "Bounded\nMean\nEBS",
    colours = c("gray98","#F5C9A9","#F2A26B","#F87E2E")
  ) +
  theme_minimal() +
  theme(
    text = element_text(family="Helvetica", color = "black", size = 7),
    axis.text.x = element_text(angle=90, size=7, vjust=0.5, hjust=1, color = "black"),
    axis.text.y = element_blank(),
    axis.title = element_text(size=7),
    legend.title = element_text(size=7),
    legend.text = element_text(size=7),
    strip.text = element_text(size=7),
    axis.ticks.y = element_blank(),
    legend.position = "none",
    panel.grid = element_blank(),
    panel.background = element_blank(),
    panel.border = element_rect(color="black", fill=NA),
    strip.text.y.left = element_text(angle=0)
  ) +
  xlab("") +
  ylab("VirScan Peptide") + 
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = "Poliovirus type",
    hjust = 1.1,
    vjust = 1.5,
    angle = -90,
    size = 8
  )

Fig_S12_A <- ggdraw() +
  draw_plot(Fig_S12_A) +
  draw_label(
    "Poliovirus",
    x = 0.99,
    y = 0.5,
    angle = -90,
    fontfamily = "Helvetica",
    size = 7
  )

Fig_S12_A
```

# Rotavirus
```{r}
# Identify vaccinated donors and prep data for virus
temp_donors <- Vaccination_Data %>%
  filter(species %in% c("Rotavirus A","Rotavirus B")) %>%
  select(donor, cohort, Sex, Age_Years, Age_Months, Pathogen) %>%
  distinct() %>%
  group_by(donor) %>%
  summarise(n_vaccination = n(), .groups="drop") %>%
  filter(n_vaccination > 1) %>%
  pull(donor) %>%
  unique()

temp_samples <- combined_metadata_dist %>% filter(donor %in% temp_donors)

# keep only donors with at least 3 sampled timepoints
temp_donors <- temp_samples %>%
  distinct(donor, identifier) %>%          # important: true sample count, not duplicated rows
  count(donor, name = "n_samples") %>%
  filter(n_samples >= 3) %>%
  pull(donor) %>%
  unique()

# re-subset samples after filtering donors
temp_samples <- combined_metadata_dist %>% filter(donor %in% temp_donors)

temp_peptides <- Viral_Annotations %>%
  filter(species %in% c("Rotavirus A","Rotavirus B")) %>%
  filter(grepl("Human", Organism))

temp_vaccination <- Vaccination_Data %>%
  filter(donor %in% temp_donors, species %in% c("Rotavirus A","Rotavirus B")) %>%
  select(-c(species, Vaccine_Name_English)) %>%
  distinct() %>%
  group_by(donor) %>%
  arrange(Age_Years) %>%
  mutate(vaccine_dose = as.factor(row_number())) %>%
  ungroup()

temp_consec_zs <- Birth_Consec_Zs %>% filter(donor %in% temp_donors, id %in% temp_peptides$id)

temp_zs <- all_zs_combined %>%
  select(c(id, all_of(temp_samples$identifier))) %>%
  filter(id %in% temp_peptides$id) %>%
  bound_Z(upper = 15) %>%
  left_join(Viral_Annotations %>% select(c(id, Entry, `Entry name`, species, `Protein names`, Organism))) %>%
  pivot_longer(-c(id, Entry, `Entry name`, species, `Protein names`, Organism),
               names_to = "identifier", values_to = "Z_Bound") %>%
  mutate(hit = ifelse(Z_Bound >= 3.5, 1, 0)) %>%
  left_join(temp_samples %>% select(c(identifier, Age_Years, donor)))

# Key timepoints: Birth + Pre (closest before Dose 1) + first sample after each available dose

# Dose 1 age per donor (first recorded vaccination age)
dose1_age <- temp_vaccination %>%
  group_by(donor) %>%
  slice_min(order_by = as.numeric(as.character(vaccine_dose)), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(donor, dose1_age = Age_Years)

# Birth sample
birth <- temp_samples %>%
  filter(Age_Years == 0) %>%
  group_by(donor) %>%
  slice_min(Age_Years, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(dose = "Birth") %>%
  select(donor, identifier, Age_Years, dose)

# Pre-vax: closest sample before Dose 1, require >= 6 months (swap to 9/12 if desired)
pre <- temp_samples %>%
  left_join(dose1_age, by = "donor") %>%
  filter(Age_Years < dose1_age) %>%
  mutate(dist_from_dose1 = abs(Age_Years - dose1_age)) %>%
  group_by(donor) %>%
  slice_min(dist_from_dose1, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  filter(Age_Years >= (6/12)) %>%   # >= 6 months
  mutate(dose = "Pre") %>%
  select(donor, identifier, Age_Years, dose)

# All post-dose selections: for each donor x dose number, choose closest sample after that dose
dose_ages <- temp_vaccination %>%
  mutate(dose_num = as.integer(as.character(vaccine_dose))) %>%
  select(donor, dose_num, dose_age = Age_Years)

post_candidates <- temp_samples %>%
  select(donor, identifier, Age_Years) %>%
  inner_join(dose_ages, by = "donor") %>%
  filter(Age_Years > dose_age) %>%
  mutate(dist_after = Age_Years - dose_age) %>%
  group_by(donor, dose_num) %>%
  slice_min(dist_after, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(dose = paste0("Dose ", dose_num)) %>%
  select(donor, identifier, Age_Years, dose)

key_metadata_raw <- bind_rows(birth, pre, post_candidates) %>%
  mutate(dose_num = case_when(
    dose == "Birth" ~ 0L,
    dose == "Pre"   ~ 0L,   # keep Pre near Birth; we’ll order via factor below
    str_detect(dose, "^Dose ") ~ as.integer(str_extract(dose, "\\d+")),
    TRUE ~ NA_integer_
  ))

# Order levels dynamically based on what exists in your data
max_dose <- key_metadata_raw %>% summarise(m = max(dose_num, na.rm = TRUE)) %>% pull(m)
dose_levels <- c("Birth", "Pre", paste0("Dose ", seq_len(max_dose)))

key_metadata_raw <- key_metadata_raw %>%
  mutate(dose = factor(dose, levels = dose_levels))

# If same identifier is chosen for multiple post-doses, keep only the highest label
key_metadata <- key_metadata_raw %>%
  group_by(donor, identifier) %>%
  slice_max(order_by = as.integer(dose), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(donor, identifier, Age_Years, dose)

# keep only donors with at least 3 sampled timepoints
temp_donors <- key_metadata %>%
  distinct(donor, identifier) %>%          # count unique plotted samples per donor
  count(donor, name = "n_key_samples") %>%
  filter(n_key_samples >= 3) %>%
  pull(donor) %>%
  unique()

# re-subset samples after filtering donors
temp_samples <- temp_samples %>% filter(donor %in% temp_donors)
key_metadata <- key_metadata %>% filter(donor %in% temp_donors)

# Join for plotting
key_zs <- temp_zs %>%
  filter(identifier %in% key_metadata$identifier) %>%
  left_join(key_metadata %>% select(identifier, dose), by = "identifier") %>%
  mutate(dose = factor(dose, levels = dose_levels))

# Donor-specific x-axis labels: "Dose (age y)"
dose_labels_rota <- key_metadata %>%
  group_by(donor, dose) %>%
  summarise(age = round(mean(Age_Years), 1), .groups = "drop") %>%
  mutate(
    dose = factor(dose, levels = dose_levels),
    dose_lab = paste0(as.character(dose), " (", age, "y)"),
    x_key = paste(donor, dose, sep = "___")
  ) %>%
  arrange(donor, dose)

key_zs2 <- key_zs %>%
  left_join(dose_labels_rota %>% select(donor, dose, dose_lab, x_key),
            by = c("donor", "dose")) %>%
  mutate(x_key = factor(x_key, levels = unique(dose_labels_rota$x_key)))

# Plot
temp_peps <- key_zs2 %>%
  group_by(id) %>%
  summarise(n_hits = sum(hit), .groups = "drop") %>%
  filter(n_hits > 0)

key_zs2 <- key_zs2 %>%
  mutate(
    RotaType = case_when(
      species == "Rotavirus A" ~ "A",
      species == "Rotavirus B" ~ "B",
      TRUE ~ species
    )
  )

rota_plot <- ggplot(
  key_zs2 %>% filter(id %in% temp_peps$id),
  aes(x = x_key, y = as.factor(id), fill = Z_Bound)
) +
  geom_tile() +
  facet_grid(RotaType ~ donor, scales = "free", space = "free") +
  scale_x_discrete(labels = setNames(dose_labels_rota$dose_lab, dose_labels_rota$x_key)) +
  scale_fill_gradientn(
    "Bounded Mean EBS",
    colours = c("gray98","#FAD0E9","#F58ACF","#FF36B0")
  ) +
  ylab("VirScan Peptide") +
  theme_minimal() +
  theme(
    text = element_text(family = "Helvetica",color = "black", size = 7),
    axis.text.x = element_text(angle = 90, size = 7, vjust = 0.5, hjust = 1, color = "black"),
    axis.text.y = element_blank(),
    axis.title = element_text(size = 7),
    legend.title = element_text(size = 7),
    legend.text = element_text(size = 7),
    strip.text = element_text(size = 7),
    axis.ticks.y = element_blank(),
    legend.position = "none",
    panel.grid = element_blank(),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    strip.text.y.left = element_text(angle = 0)
  ) +
  xlab("")

Fig_S12_B <- ggdraw() +
  draw_plot(rota_plot) +
  draw_label(
    "Rotavirus",
    x = 0.99,
    y = 0.5,
    angle = -90,
    fontfamily = "Helvetica",
    size = 7
  )

Fig_S12_B
```

# Create legend and combine figure S12
```{r, fig.height=7,fig.width=7}
maxZ<- 15
legend_breaks <- c(0, 5, 10, maxZ)
legend_labels <- c("0", "5", "10", as.character(maxZ))

legend_polio <- ggplot(legend_df, aes(x = z, y = y, fill = meas_hex)) +
  geom_tile(height = 0.9) +
  annotate("text", x = maxZ * 0.5, y = 1, label = "Polio",
           family = "Helvetica", size = 2.5,  color = "black") +
  scale_fill_identity() +
  scale_x_continuous(breaks = legend_breaks, labels = legend_labels, expand = c(0, 0)) +
  theme_minimal() +
  theme(
    text = element_text(family = "Helvetica", color = "black", size = 7),
    axis.title.x = element_blank(),
    axis.text.x  = element_text(size = 7, color = "black"),
    axis.ticks.x = element_line(linewidth = 0.3, color = "black"),
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid   = element_blank(),
    plot.margin  = margin(t = 0, r = 4, b = 0, l = 4)
  )

legend_rota <- ggplot(legend_df, aes(x = z, y = y, fill = rub_hex)) +
  geom_tile(height = 0.9) +
  annotate("text", x = maxZ * 0.5, y = 1, label = "Rotavirus",
           family = "Helvetica", size = 2.5,  color = "black") +
  scale_fill_identity() +
  scale_x_continuous(breaks = legend_breaks, labels = legend_labels, expand = c(0, 0)) +
  theme_minimal() +
  theme(
    text = element_text(family = "Helvetica", color = "black", size = 7),
    axis.title.x = element_blank(),
    axis.text.x  = element_text(size = 7, color = "black"),
    axis.ticks.x = element_line(linewidth = 0.3, color = "black"),
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid   = element_blank(),
    plot.margin  = margin(t = 0, r = 4, b = 0, l = 4)
  )

legend_title <- ggplot() +
  annotate("text", x = 0.5, y = 0.5, label = "Bounded Mean EBS",
           family = "Helvetica", size = 3, fontface = "plain", color = "black") +
  theme_void() +
  theme(plot.margin = margin(t = 0, r = 0, b = 0, l = 0))

top_legends <- cowplot::plot_grid(
  legend_polio + theme(plot.margin = margin(0, 5, 0, 0)),
  legend_rota + theme(plot.margin = margin(0, 0, 0, 5)),
  ncol = 2,
  rel_widths = c(1, 1)
)

panel_top <- cowplot::plot_grid(
  legend_title,
  top_legends,
  ncol = 1,
  rel_heights = c(0.35, 1)
)

ggarrange(
  ggarrange(
    ggplot(NULL) + theme_void(),
    panel_top,
    ggplot(NULL) + theme_void(),
    widths = c(0.75, 18, 0.75),
    nrow = 1
  ),
  Fig_S12_A,
  Fig_S12_B,
  labels = c("a", "", "b"),
  ncol = 1,
  font.label = list(family = "Helvetica", size = 8),
  heights = c(0.12, 1, 0.8)
)
ggsave(file=paste0(folder_path_save, "/AVARS_Figure_S12_Final.pdf"))
dev.off()
```
