---
title: "Figure_4_Figure_S4"
output: html_notebook
---

```{r}
# Load packages
easypackages::libraries(c("readr","readxl","dplyr","tidyr","ggplot2","ggpubr","UpSetR","cowplot","grid","ggplotify","data.table","plotROC","pROC","lme4", "splines","emmeans","tidytext","here"))

# insert path to folders
# Define paths relative to the root of the cloned GitHub repository
directory_path <- here::here()
folder_path <- paste0(directory_path, "/Supplemental Tables/") # folder containing downloaded data sets
intermediate_folder_path <- paste0(directory_path, "/Intermediate Files Used in Plots/") # path containing intermediate outputs
folder_path_save <- paste0(directory_path, "/Complete Figures/")# define folder path to save images to
folder_path_code <- paste0(directory_path, "/Functions and General Analysis Code/") # change this to folder containing downloaded functions and code

# Load Supplemental Data
combined_metadata_dist <- read_csv(paste0(folder_path,"Data Table S1.csv"))
all_zs_combined <- read_csv(paste0(folder_path,"Data Table S2.csv"))
all_hits_combined <- read_csv(paste0(folder_path,"Data Table S3.csv"))
Viral_Annotations <- read_csv(paste0(folder_path,"Data Table S4.csv"))

# Load Intermediate data
Z_fam_cor_combined <- read_csv(paste0(intermediate_folder_path,"Viral_Family_Autocorrelation_Relative_to_initial_samples_by_age.csv"))  # Data frame used was generated in the script "Generate Viral Family Correlation Trajectories.R"
adult_Consec_Zs <- read_csv(paste0(intermediate_folder_path,"Adult_Consec_Zs.csv")) # Data frame used was generated in the script "Generate Consecutive Z Score Files.R"
ped_5_Zs <- read_csv(paste0(intermediate_folder_path,"Ped_5_apart_Zs.csv")) # Data frame used was generated in the script "Generate Consecutive Z Score Files.R"

pal1<-c("#F87E2E", "#FFEC33", "#1FAA79", "#14C4F8", "#0077C8", "#9870F5", "#D53EE1", "#FF36B0", "#D72638", "#6E6E6E")
source(paste0(folder_path_code,"Immunoprint functions.R"))

```

# Prepare Data for Figure 4 A-B
```{r}
set.seed(10)

# Figure 4 A-B use D579 as an example
temp_donors<-"D579"
temp_meta<-combined_metadata_dist %>% filter(donor %in% temp_donors)
temp_zs<-all_zs_combined %>% select(c(id,temp_meta$identifier))

# Use trajectory functions to prepare data for 4 A-B
input1 <- generate_heat_frame(temp_zs,Viral_Annotations, temp_meta)
input2 <- generate_trajectory_frame(temp_zs, Viral_Annotations, temp_meta)

# Order according to age
temp_order<-input1 %>% select(c(samp_label,Age_Years)) %>% distinct() %>% arrange(Age_Years)
temp <- input1  %>% group_by(family_label,`Protein names`,donor) %>% summarise(sum_adj=sum(adjZ)) %>% filter(sum_adj>3.5) # limit to proteins that have been targeted by the donor
temp_heat<-temp %>% left_join(input1)
temp<-temp_heat %>% group_by(id,donor) %>% summarise(max_adj_pep=max(adjZ)) %>% filter(max_adj_pep>3.5) # limit to peptides that have been targeted in the donor
input1<-temp %>% left_join(temp_heat)
rm(temp,temp_heat)

# Prepare correlation data for panel B
Z_fam_cor_long <- input2 %>% select(-c(label))%>% pivot_longer(-c(sample1,sample2,donor1,donor2,Age_Years1,Age_Years2,samp1,samp2,years_between),names_to = "family_label",values_to = "fam_cor")
```

# Generate Figure 4 A-B
```{r,fig.width=7,fig.height=6}
# Panel A
input1_ord <- input1 %>%
  group_by(family_label) %>%
  arrange(species, as.numeric(id), .by_group = TRUE) %>%
  mutate(row_rank = row_number()) %>%
  ungroup() %>%
  mutate(id_reord = reorder_within(id, row_rank, family_label))

Fig_4_A <- ggplot(input1_ord) +
  geom_tile(aes(
    x = factor(samp_label, levels = c(temp_order$samp_label)),
    y = id_reord,
    fill = adjZ
  )) +
  facet_wrap(
    ~ factor(family_label, levels = c("Herpesviridae","Picornaviridae","Orthomyxoviridae",
                                     "Pneumoviridae","Adenoviridae","Coronaviridae","Paramyxoviridae")),
    scales = "free_y", nrow = 1
  ) +
  scale_y_reordered() +
  theme_minimal() +
  theme(
    text = element_text(family="Helvetica", size=7),
    axis.text.y = element_blank(),
    axis.text.x = element_text(angle = 90, size=7),
    axis.ticks = element_blank(),
    strip.text.y = element_blank(),
    legend.position = "top",
    legend.key.height = unit(0.2, "cm"), 
    legend.key.width = unit(1.0, "cm"),
    legend.margin = margin(b = -5, unit = "pt"), 
    legend.box.margin = margin(t = 0, b = -5, unit = "pt"),
    legend.spacing.x = unit(0.1, 'cm'),
    legend.spacing.y = unit(0, 'cm'),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size=7),
    strip.text.x = element_text(size=7),
    legend.title = element_text(size=7),
    panel.grid = element_blank(),
    panel.background = element_blank(),
    panel.spacing = unit(0.07, "lines")
  ) +
  scale_fill_gradientn(
    "Bounded\nMean EBS",
    colours = c("white","gray90",pal1[9],"red3"),
    breaks = c(0,5,10,15)
  ) +
  ylab("VirScan\nPeptide")
Fig_4_A

# Panel B
Fig_4_B<-ggplot(Z_fam_cor_long, aes(x=years_between,alpha=0.7)) +
  geom_line(aes(y=fam_cor,color=family_label),size=1) + 
  ylab("Correlation") + 
  facet_wrap(~factor(family_label,levels = c("Herpesviridae", "Picornaviridae", "Orthomyxoviridae", "Pneumoviridae", "Adenoviridae", "Coronaviridae", "Paramyxoviridae")), nrow=1) +
  theme_minimal() +
  scale_color_manual(name="Viral\nfamily",values = pal1[c(1,3:8)], breaks = c("Herpesviridae", "Picornaviridae", "Orthomyxoviridae", "Pneumoviridae", "Adenoviridae", "Coronaviridae", "Paramyxoviridae"), guide=guide_legend(override.aes = list(alpha = 0.6),reverse = TRUE)) +
  theme(text=element_text(family="Helvetica",size=7),
        legend.position = "none",
        strip.text.x = element_text(size=7),
        axis.title = element_text(size=7),
        axis.text = element_text(size = 7, color = "black"),
        axis.text.x = element_text(angle=45),
        panel.spacing = unit(0.25, "lines")) + 
  ylim(c(-0.01,1)) +
  xlab("Sampling interval (years)") +
  scale_x_continuous(breaks = c(0, 5, 10))
Fig_4_B
```
# Generate Data for Figure 4 C
Data frame used was generated in the script "Generate Viral Family Correlation Trajectories.R"
# Quantitatively compare stability across families within age group using a semi-parametric mixed model

```{r}
Z_fam_cor_combined_long <- Z_fam_cor_combined %>% select(-c(label))%>% pivot_longer(-c(sample1,sample2,donor1,donor2,Age_Years1,Age_Years2,samp1,samp2,years_between,facet_label),names_to = "family_label",values_to = "fam_cor") %>% 
  mutate(
  facet_label = stringr::str_replace(
    facet_label,
    "\\n\\((\\d+)\\)$",
    "\n(n = \\1)"
  )
)
## Fisher Z Transform correlations for model
Z_fam_cor_combined_long <- Z_fam_cor_combined_long %>%
  mutate(
    z_cor = atanh(pmin(pmax(fam_cor, -0.9999), 0.9999)) # avoid ±1
  )
# Nonlinear trajectory with splines
fit_spline <- lmer(
  z_cor ~ ns(years_between, df = 3) * family_label * facet_label + (1 | donor1),
  data = Z_fam_cor_combined_long %>% filter(years_between <= 12)
)

# Compare between families at fixed time points
emm_time <- emmeans(
  fit_spline,
  ~ family_label | facet_label * years_between,
  at = list(years_between =c(0,5,10)),pbkrtest.limit = 4256
)

# Pairwise comparisons of families within each age group & timepoint
pairwise_time <- contrast(emm_time, method = "pairwise", by = c("facet_label", "years_between"))

summary(pairwise_time)

# Transform back to original correlation scale
emm_time_corr <- emm_time %>%
  as.data.frame() %>%
  mutate(
    est_r   = tanh(emmean),
    lower_r = tanh(lower.CL),
    upper_r = tanh(upper.CL),
    # ensure consistent family order for facets
    family_label = factor(
      family_label,
      levels = c(
        "Herpesviridae", "Picornaviridae",
        "Orthomyxoviridae", "Pneumoviridae", "Adenoviridae",
        "Coronaviridae", "Paramyxoviridae"
      )
    )
  )

emm_time_corr %>% filter(years_between !=0)
```
# Compare correlations at 5 years

```{r}
# Adult ranking
emm_time_corr %>% filter(years_between == 5,facet_label=="Adults\n(n = 96)") %>%
  arrange(desc(est_r))


# Statistical comparison
emm_5yr <- subset(emm_time, years_between == 5) 
emm_5yr <- subset(emm_5yr,facet_label=="Adults\n(n = 96)")
emm_5yr

pairwise_fam_5yr <- emmeans::contrast(emm_5yr, method = "pairwise")

summary(pairwise_fam_5yr, adjust = "BH") %>% arrange(p.value)   # or adjust = "tukey"
```

# Plot 5 Year R
```{r}
# Define family set + colors as in Fig 4B
fam_levels <- c(
  "Herpesviridae",
  "Picornaviridae",
  "Orthomyxoviridae",
  "Pneumoviridae",
  "Adenoviridae",
  "Coronaviridae",
  "Paramyxoviridae"
)

family_cols <- setNames(
  pal1[c(1, 3:8)],
  fam_levels
)

#  Compute ordering from Adults (96) at 5 years
fam_order_adults <- emm_time_corr %>%
  dplyr::filter(years_between == 5, facet_label == "Adults\n(n = 96)") %>%
  dplyr::arrange(dplyr::desc(est_r)) %>%
  dplyr::pull(family_label) %>%
  as.character()

# keep only the families you expect, and keep the expected order if anything is missing
fam_order_adults <- fam_order_adults[fam_order_adults %in% fam_levels]
if (length(fam_order_adults) == 0) fam_order_adults <- fam_levels

# 3) Apply ordering 
emm_time_corr_ord <- emm_time_corr %>%
  dplyr::mutate(
    family_label = factor(family_label, levels = fam_order_adults)
  )

#  plot
Fig_4_D <- ggplot2::ggplot(
  emm_time_corr_ord %>% dplyr::filter(years_between == 5),
  ggplot2::aes(x = family_label, y = est_r, color = family_label)
) +
  ggplot2::geom_point(size = 2) +
  ggplot2::geom_errorbar(
    ggplot2::aes(ymin = lower_r, ymax = upper_r),
    width = 0.15,
    linewidth = 0.6
  ) +
  ggplot2::coord_flip() +
  ggplot2::facet_wrap(~ facet_label) +
  ggplot2::scale_color_manual(values = family_cols, breaks = fam_levels) +
  ggplot2::theme_minimal() +
  ggplot2::theme(
    text = ggplot2::element_text(family = "Helvetica",size=7),
    legend.position = "none",
    strip.text.x = ggplot2::element_text(size = 7),
    axis.title = ggplot2::element_text(size=7),
    axis.text = ggplot2::element_text(size=7),
    panel.spacing = grid::unit(0.25, "lines"),
    panel.border = ggplot2::element_rect(fill = NA, color = "black"),
    panel.grid.y = ggplot2::element_blank()
  ) +
  ggplot2::labs(
    y = "Estimated 5-year correlation",
    x = "Viral family"
  ) 

Fig_4_D

```


# Generate Figure 4 C
```{r}
# Plot
family_levels <- c(
  "Herpesviridae", "Picornaviridae",
  "Orthomyxoviridae", "Pneumoviridae", "Adenoviridae",
  "Coronaviridae", "Paramyxoviridae"
)

Z_fam_cor_combined_long <- Z_fam_cor_combined_long %>%
  mutate(
    family_label = factor(family_label, levels = family_levels)
  )

emm_time_corr <- emm_time_corr %>%
  mutate(
    family_label = factor(family_label, levels = family_levels)
  )

Fig_4_C_alt <- ggplot() + geom_line(
    data = Z_fam_cor_combined_long %>% filter(years_between <= 12),
    aes(
      x     = years_between,
      y     = fam_cor,
      color = family_label,
      group = donor1
    ),
    alpha = 0.3,
    size  = 0.5
  ) +
  # model-based CI ribbon
  geom_ribbon(
    data  = emm_time_corr,
    aes(
      x     = years_between,
      ymin  = lower_r,
      ymax  = upper_r,
      group = interaction(facet_label, family_label)
    ),
    inherit.aes = FALSE,
    fill        = "grey50",
    alpha       = 0.6
  ) +
  # model-based mean curve
  geom_line(
    data = emm_time_corr,
    aes(
      x     = years_between,
      y     = est_r,
      color = family_label,
      group = family_label
    ),
    inherit.aes = FALSE,
    size        = 1.0
  ) +
  ylab("Correlation") +
  facet_grid(facet_label ~ family_label) +
  theme_minimal() +
  scale_color_manual(
    name   = "Viral\nfamily",
    values = pal1[c(1,3:8)],
    breaks = family_levels,
    guide  = guide_legend(
      override.aes = list(alpha = 0.6),
      reverse = TRUE
    )
  ) +
  theme(
    text            = element_text(family = "Helvetica",size=7),
    legend.position = "none",
    strip.text.x    = element_text(size = 7),
    strip.text.y    = element_text(size = 7),
    axis.title      = element_text(size = 7),
    axis.text       = element_text(size = 7, color = "black"),
    axis.text.x     = element_text(angle = 45, hjust = 1),
    panel.spacing   = unit(0.5, "lines")
  ) +
  xlim(c(0, 12)) +
  xlab("Sampling interval (years)") +
  scale_x_continuous(breaks = c(0, 5, 10))

Fig_4_C_alt
```


# Figure 4 Combined
```{r,fig.width=7,fig.height=8}
ggarrange(ggarrange(Fig_4_A + theme(legend.text = element_text(size=6), axis.text.x=element_text(size=7, color="black")),ggplot(NULL)+theme_void(),widths = c(14,0.75)),ggarrange(Fig_4_B+theme(axis.text.x = element_text(angle = 45, size=7)),ggplot(NULL)+theme_void(),widths = c(14,0.75)),Fig_4_C_alt,ggarrange(Fig_4_D, ggplot(NULL)+theme_void(), nrow=1, widths = c(14,0.2)) ,nrow = 4,heights = c(1.1,0.70,2.05,1),labels=c("a","b","c","d"),font.label = list(family="Helvetica",size=8))
ggsave(file=paste0(folder_path_save, "/AVARS_Figure_4_Final.pdf"))
dev.off()
```


# Address different number of peptides per family
## Do analysis on the viral families with greater library representation (Herpesviruses ~17000, Other~25000, Orthomyxoviriae ~7500)
 family_label     n_entries n_peptides
1 Adenoviridae           575       4838
2 Coronaviridae           70       2841
3 Herpesviridae         1571      17072
4 Orthomyxoviridae      1165       7535
5 Paramyxoviridae        101       1286
6 Picornaviridae         137       5069
7 Pneumoviridae          199       1099

### Sample varying numbers of peptides 
Number of peptides ranges from 1099-17072
sample: 1000, 2000, 5,000, 7500, 10,000, 15,000, 17,000 from Herpesviridae
sample: 1000, 2000, 5,000, 7500 from Orthmyxoviridae
sample: 1000, 2000, 5,000 from Picornaviridae
```{r}
downsampled_corr <- read.csv(file=paste0(intermediate_folder_path, "Down_sampled_viral_family_correlation_adult.csv")) # Generated in "Generate Viral Family Correlation Trajectories.R"

downsampled_fits <- summarize_downsampled_by_model(
  downsampled_corr,
  time_points = c(0, 2, 5)
)

# keep 5y and 10y only, and label them for faceting
downsampled_plot <- downsampled_fits %>%
  filter(years_between %in% c(2, 5)) %>%
  mutate(
    n_peptides_sampled = as.numeric(as.character(n_peptides_sampled)),
    time_label = factor(
      years_between,
      levels = c(2, 5),
      labels = c("2-year sampling interval", "5-year sampling interval")
    )
  ) %>%
  filter(!is.na(n_peptides_sampled), !is.na(est_r))

# summary (median + 95% interval) WITHIN each time_label
summ_plot <- downsampled_plot %>%
  group_by(time_label, family_label, n_peptides_sampled) %>%
  summarize(
    med = median(est_r, na.rm = TRUE),
    lo  = quantile(est_r, 0.025, na.rm = TRUE),
    hi  = quantile(est_r, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

Fig_S_4_A <- ggplot(downsampled_plot) +
  # individual downsampling iterations
  geom_line(
    aes(
      x     = n_peptides_sampled,
      y     = est_r,
      group = iter
    ),
    alpha = 0.12
  ) +

  # summary ribbon (95% interval across iterations)
  geom_ribbon(
    data = summ_plot,
    aes(
      x     = n_peptides_sampled,
      ymin  = lo,
      ymax  = hi,
      group = 1
    ),
    inherit.aes = FALSE,
    alpha = 0.3,
    fill  = pal1[6]
  ) +

  # summary line (median)
  geom_line(
    data = summ_plot,
    aes(
      x     = n_peptides_sampled,
      y     = med,
      group = 1
    ),
    inherit.aes = FALSE,
    size  = 0.8,
    color = pal1[6]
  ) +

  # summary dots (median)
  geom_point(
    data = summ_plot,
    aes(
      x     = n_peptides_sampled,
      y     = med,
      group = 1
    ),
    inherit.aes = FALSE,
    size  = 1.1,
    color = pal1[6]
  ) +

  # row = time, columns = family
  facet_grid(time_label ~ family_label) +

  theme_minimal() +
  xlab("Number of peptides sampled") +
  ylab("Model-estimated correlation") +
  ylim(c(0, 1)) +

  theme(
    text            = element_text(family = "Helvetica", size=7),
    legend.position = "none",
    strip.text.x    = element_text(size = 7),
    strip.text.y    = element_text(size = 7),
    axis.title      = element_text(size = 7),
    axis.text       = element_text(size = 7),
    panel.border    = element_rect(fill = NA, color = "black"),
    panel.grid      = element_blank()
  )
```
### Heatmap visualization
```{r}
heat_df <- downsampled_fits %>%
  filter(years_between %in% c(2, 5)) %>%
  mutate(
    n_peptides_sampled = as.numeric(as.character(n_peptides_sampled)),
    time_label = factor(
      years_between,
      levels = c(2, 5),
      labels = c("2 years", "5 years")
    )
  ) %>%
  group_by(time_label, family_label, n_peptides_sampled) %>%
  summarize(
    med_est_r = median(est_r, na.rm = TRUE),
    .groups = "drop"
  )
family_levels_2 <- c("Herpesviridae","Orthomyxoviridae","Picornaviridae")
heat_df <- heat_df %>%
  mutate(
    family_label = factor(family_label, levels = family_levels_2)
  )

Fig_S_4_B <- ggplot(heat_df, aes(x = as.factor(n_peptides_sampled), y = family_label, fill = med_est_r)) +
  geom_tile(color = "black", size = 0.3) +
  geom_text(aes(label = sprintf("%.2f", med_est_r)), size = 2.5) +
  facet_grid(~ time_label) +
  theme_minimal() +
  theme(
    text = element_text(family = "Helvetica", color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    strip.text.x    = element_text(size = 7),
    strip.text.y    = element_text(size = 7),
    axis.title      = element_text(size = 7),
    axis.text       = element_text(size = 7),
  ) +
  labs(x="Number of peptides sampled", y="Viral family", fill="Median est_r") +
  coord_fixed() + 
  scale_fill_gradientn(colours = c("white",pal1[4],pal1[8]),limits=c(0,1), name = "Median\nestimated r") 
```

### Combine Panels
```{r, fig.width=7,fig.height=6}
ggarrange(Fig_S_4_A, Fig_S_4_B + theme(legend.title = element_text(size=7),legend.text = element_text(size=7)), labels=c("a","b"),font.label = list(family="Helvetica",size=8),nrow=2, heights = c(2.5,1.4))
ggsave(file=paste0(folder_path_save, "/AVARS_Figure_S4_Final.pdf"))
dev.off()
```
  









